---
title: "EDA 65 — Missing Data Mechanism: Little's MCAR Test & Missingness Visualization"
subtitle: "Systematic assessment of missing-data patterns across the Andersen Model domains"
author: "Analysis Team"
date: "2026-05-23"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: true
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This EDA conducts a systematic missing-data assessment for the CCHS analytic dataset.
It addresses the request from Marc-André Blanchette (2026-05-15 review) to evaluate the
missing-data mechanism using Little's MCAR test and to visualize variable-level
missingness patterns.

**Research Questions:**

1. What is the missingness rate for each analysis variable, and do variables cluster by
   missingness level?
2. What are the most common missingness *patterns* (combinations of jointly-missing variables)?
3. Is missingness completely at random (MCAR) within each Andersen model domain, or is
   there evidence of structure (MAR/MNAR)?
4. Does missingness concentrate in specific age groups (evidence of age-gated questions)?
5. Does missingness differ between the two CCHS cycles?

**Method:** Little's MCAR test (Little, 1988) tests the null hypothesis that missing data
are Missing Completely At Random. A significant result (p < 0.05) indicates that the
pattern of missingness depends on observed values — i.e., the data are at least MAR.

> **Reference:** Little, R. J. A. (1988). A test of missing completely at random for
> multivariate data with missing values. *Journal of the American Statistical
> Association*, 83(404), 1198–1202.

Following the dual-file development pattern, all analytical code lives in
`analysis/eda-65/eda-65.R`. This document sources those chunks for presentation.

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = "../../")
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results   = "show",
  message   = FALSE,
  warning   = FALSE,
  comment   = NA,
  tidy      = FALSE,
  out.width = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-65/eda-65.R")
```

# Environment

> Reviews the components of the working environment. Non-technical readers may skip this section.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
```

```{r}
#| label: load-sources
#| code-summary: Project-level scripts
#| echo: true
#| results: hide
```

```{r}
#| label: declare-globals
#| code-summary: Paths and variable definitions
#| echo: true
#| results: hide
```

```{r}
#| label: declare-functions
#| code-summary: Analysis-specific helper functions
#| echo: true
```

# Data

```{r}
#| label: load-data
#| code-summary: Load analytical parquet
#| cache: true
```

```{r}
#| label: data-context-tables
#| code-summary: Data provenance
#| echo: false
```

---

# Graph Family G1: Variable-Level Missingness

<!-- Research question: What is the missingness rate for each variable? -->

```{r}
#| label: g1-data-prep
#| code-summary: Calculate per-variable missingness
#| code-fold: true
#| cache: true
```

```{r}
#| label: g1
#| code-summary: Missingness rate by variable (bar chart)
#| code-fold: true
#| fig-cap: "Per-variable missingness rate coloured by Andersen model domain"
#| fig-height: 9
```

```{r}
#| label: g11
#| code-summary: Missingness heatmap (subsample)
#| code-fold: true
#| fig-cap: "Missingness pattern visualisation (5,000-row random subsample)"
#| fig-height: 7
```

---

# Graph Family G2: Missingness Patterns

<!-- Research question: What are the most common combinations of missing variables? -->

```{r}
#| label: g2-data-prep
#| code-summary: Calculate missingness patterns
#| code-fold: true
#| cache: true
```

```{r}
#| label: g2
#| code-summary: Number of missing variables per respondent
#| code-fold: true
#| fig-cap: "Distribution of missingness burden across the sample"
```

```{r}
#| label: g21
#| code-summary: Top 15 missingness patterns (upset-style)
#| code-fold: true
#| fig-cap: "Most frequent combinations of jointly-missing variables"
#| fig-height: 7
```

---

# Graph Family G3: Little's MCAR Test

> **Little's MCAR test** (1988) evaluates the null hypothesis that missingness is
> completely at random. The test is run separately for each Andersen model domain and
> for all predictors combined. A significant p-value (p < 0.05) indicates that
> missingness is *not* MCAR — it depends on observed values (at minimum MAR).

```{r}
#| label: g3-data-prep
#| code-summary: Run Little's MCAR test by domain
#| code-fold: true
#| cache: true
```

```{r}
#| label: g3
#| code-summary: MCAR test summary plot
#| code-fold: true
#| fig-cap: "Little's MCAR test by domain — bars above the dashed line reject MCAR"
```

**Interpretation guide:**

- If **p < 0.05** (bar above dashed line): Reject MCAR → missingness has structure.
  This is expected for age-gated variables (COPD, BMI) where younger respondents are
  systematically missing.
- If **p ≥ 0.05** (bar below line): Consistent with MCAR → missingness appears random.
- The test cannot distinguish MAR from MNAR — only that MCAR is violated.

---

# Graph Family G4: Missingness by Age Group

> **Context**: If missingness is concentrated in the youngest age group (15–24), this
> confirms age-gated survey design rather than data quality problems.

```{r}
#| label: g4-data-prep
#| code-summary: Missingness by age group
#| code-fold: true
#| cache: true
```

```{r}
#| label: g4
#| code-summary: Heatmap of missingness by age group
#| code-fold: true
#| fig-cap: "Variables with high missingness in the 15-24 group are age-gated questions"
#| fig-height: 8
```

---

# Graph Family G5: Missingness by Cycle

```{r}
#| label: g5-data-prep
#| code-summary: Missingness by CCHS cycle
#| code-fold: true
#| cache: true
```

```{r}
#| label: g5
#| code-summary: Paired dot-plot of missingness by cycle
#| code-fold: true
#| fig-cap: "Comparing % missing between 2010-2011 and 2013-2014 cycles"
#| fig-height: 8
```

---

# Summary

```{r}
#| label: save-results
#| echo: false
#| results: hide
```

```{r}
#| echo: false
d_local <- readRDS("analysis/eda-65/data-local/eda-65.rds")
knitr::kable(
  d_local$mcar_results %>% select(block, statistic, df, p_value, note),
  col.names = c("Domain Block", "Chi-sq", "df", "p-value", "Interpretation"),
  digits    = c(0, 1, 0, 4, 0),
  caption   = "Little's MCAR Test Results by Variable Domain"
)
```

---

```{r}
#| label: session-info
#| echo: false
#| results: hide
report_render_stop_time <- Sys.time()
cat(
  "Report rendered at:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
  "\nElapsed:", round(difftime(report_render_stop_time, report_render_start_time, units = "secs"), 1),
  "seconds\n"
)
sessionInfo()
```
