---
title: "EDA 3 - Questions 4-1 and 4-2"
subtitle: "Outcome Comparison and Weighted Descriptive Statistics"
author: "Analysis Team"
date: "2026-03-22"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This document **observes and documents** question **4-1** and question **4-2** from `stats_instructions_v3.md`, using the existing Ellis output in `data-private/derived/cchs-2-tables/`. It keeps the same reporting style used in `eda-2.qmd`: requirement framing, explicit code-based proof chunks, and concise completion callouts.

**Data genealogy for this report:**

```         
1-ferry.R  ──► cchs-1.sqlite ──► 2-ellis.R ──► cchs_analytical.parquet ──► EDA-3 (Q4-1, Q4-2)
```

This report is split into two parts:

-   **Part A — Q4-1 Outcome Comparison**: primary vs sensitive absenteeism outcomes
-   **Part B — Q4-2 Weighted Distributional Profile**: required weighted descriptive statistics and distribution checks

```{r}
#| label: knitr-root
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../')
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results = 'show',
  message = FALSE,
  warning = FALSE,
  comment = NA,
  tidy = FALSE,
  out.width = "960px",
  fig.width = 10,
  fig.height = 6,
  fig.path = 'figure-png-iso/'
)
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-3/eda-3.R")
```

# Environment

```{r}
#| label: load-packages
#| code-summary: Packages used in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Project-level functions
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Paths and constants
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Local helper functions
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

------------------------------------------------------------------------

# Part A — Question 4-1 Outcome Comparison {#part-a}

## A.1 Inputs

```{r}
#| label: load-data
#| code-summary: Load analytical dataset from Ellis output
#| code-fold: true
#| cache: true
```

::: {.callout-note icon="false"}
**Requirement — §4.1** (`stats_instructions_v3.md`, verbatim):

The primary outcome is the total number of workdays missed in the three months preceding the survey due to any health-related reason. This is a count variable derived from the LOP module by summing days absent across all reported health reasons:

- Chronic condition
- Injury
- Cold (runny nose, congestion, cough)
- Gastroenteritis
- Flu/influenza
- Other respiratory infection
- Other infectious disease
- Any other reason related to physical or mental health

The previous analysis specifies a minimum value of 1 day and a maximum of 90 days. Participants reporting zero days absent have the value 0 (not missing).

The sensitivity analysis uses days absent due to chronic conditions only. This restricted outcome variable should be constructed separately.
:::

The analytical table used is `cchs_analytical.parquet` with pooled survey weight `wts_m_pooled`.

## A.2 Data Context

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: Ellis tables used and rationale
#| code-fold: true
#| cache: true
print(data_context_tables)
print(data_context_grain)
```

```{r}
#| label: data-context-person
#| code-summary: Single-person record view
#| code-fold: true
#| cache: true
print(data_context_person)
```

```{r}
#| label: data-context-distributions
#| code-summary: Key variable distributions
#| code-fold: true
#| cache: true
print(data_context_distributions)
```

## A.3 Q4-1 Analysis

Question 4-1 is implemented as a **comparative availability analysis** between the two outcome definitions, focused on the number of observed rows and weighted coverage by outcome.

To make the construction pathway explicit, this section also anchors the outcome columns to Ferry/Ellis logic documented in `manipulation/pipeline.md`, `analysis/eda-2/eda-2.qmd`, and `analysis/eda-2/eda-2.R`.

### A.3.1 How each outcome column is built (code provenance)

From the Ellis lane (`manipulation/2-ellis.R`), the two columns are generated in `# ---- tweak-data-1-outcomes ----`:

```r
lop_vars <- c("lopg040", "lopg070", "lopg082", "lopg083",
              "lopg084", "lopg085", "lopg086", "lopg100")

days_absent_total = rowSums(across(all_of(lop_vars), ~ as.numeric(.x)), na.rm = TRUE)
outcome_all_na = rowSums(is.na(across(all_of(lop_vars)))) == length(lop_vars)
days_absent_total = if_else(outcome_all_na, 0, days_absent_total)

days_absent_chronic = as.numeric(lopg040)
```

Interpretation for this report:

- `days_absent_total` is the 8-reason sum (with all-NA rows set to structural zero).
- `days_absent_chronic` is the chronic-only component (`lopg040`).
- EDA-2 documents this same lineage at the pipeline level (Ferry -> Ellis -> `cchs_analytical.parquet`), and EDA-3 reads that Ellis output directly.

```{r}
#| label: g1-data-prep
#| code-summary: Prepare and summarize Q4-1 outcomes
#| code-fold: true
#| cache: true
```

```{r}
#| label: g1-data-prep-display
#| code-summary: Display Q4-1 row-count comparison table
#| code-fold: true
#| cache: true
print(q4_1_count_comparison)
```

```{r}
#| label: g1
#| code-summary: Row-count comparison for primary vs sensitive outcome
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-1 comparison of observed rows by outcome definition (non-missing vs missing)."
```

```{r}
#| label: g1-display
#| code-summary: Render Q4-1 row-count comparison figure
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-1 comparison of observed rows by outcome definition (non-missing vs missing)."
print(g1_outcome_row_counts)
```

```{r}
#| label: g11
#| code-summary: Weighted coverage comparison for non-missing outcomes
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-1 weighted non-missing population retained by outcome definition."
```

```{r}
#| label: g11-display
#| code-summary: Render weighted non-missing coverage comparison
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-1 weighted non-missing population retained by outcome definition."
print(g11_outcome_weighted_coverage)
```

```{r}
#| label: verify-q41-proof
#| code-summary: "§4.1 PROOF: weighted comparison table, row-count comparison, and mean-gap check"
#| code-fold: true
verify_q41_proof <- q4_1_summary %>%
  transmute(
    outcome,
    n_obs_non_missing,
    weighted_mean = round(weighted_mean, 4),
    weighted_se_mean = round(weighted_se_mean, 4)
  )

q4_1_mean_gap_proof <- q4_1_comparison %>%
  summarise(
    mean_difference_primary_minus_sensitive =
      weighted_mean[outcome == "Primary outcome: days_absent_total"] -
      weighted_mean[outcome == "Sensitive outcome: days_absent_chronic"]
  ) %>%
  mutate(
    mean_difference_primary_minus_sensitive =
      round(mean_difference_primary_minus_sensitive, 4)
  )

q4_1_count_gap_proof <- q4_1_count_gap %>%
  mutate(across(everything(), ~ round(.x, 4)))

print(verify_q41_proof)
print(q4_1_count_comparison)
print(q4_1_count_gap_proof)
print(q4_1_mean_gap_proof)
```

::: {.callout-tip icon="false"}
**§4.1 DONE — Primary vs sensitive outcomes compared** ✓

The report now focuses first on the observable row-level comparison between `days_absent_total` and `days_absent_chronic` (non-missing/missing counts + weighted non-missing coverage), with proof tables that also include the weighted mean gap check.
:::

------------------------------------------------------------------------

# Part B — Question 4-2 Weighted Statistics {#part-b}

## B.1 Requirement Framing

::: {.callout-note icon="false"}
**Requirement — §4.2** (`stats_instructions_v3.md`, verbatim):

Before modeling, verify and report the following descriptive statistics for the outcome variable:

- Mean and standard error (weighted)
- Median, Q1, Q3 (weighted)
- Variance and standard deviation (weighted)
- Frequency and proportion of zero values
- Maximum value
- Pearson dispersion statistic (variance/mean ratio)
:::

## B.2 Weighted Statistics Table

The table below is the main code-backed output for Q4-2.

```{r}
#| label: g2-data-prep
#| code-summary: Compute weighted statistics for Q4-2 requirements
#| code-fold: true
#| cache: true
```

```{r}
#| label: g2-data-prep-display
#| code-summary: Display Q4-2 weighted statistics table (reader-friendly)
#| code-fold: true
#| cache: true
q4_2_stats_pretty %>%
  mutate(`Zero proportion (weighted)` = scales::percent(`Zero proportion (weighted)`, accuracy = 0.01)) %>%
  knitr::kable(caption = "Q4-2 required weighted descriptive statistics by outcome")
```

## B.3 Distributional Checks

```{r}
#| label: g2
#| code-summary: Weighted distribution comparison for both outcomes
#| code-fold: true
#| cache: true
#| fig-cap: "Weighted histograms of absenteeism outcomes."
```

```{r}
#| label: g2-display
#| code-summary: Render weighted distribution histogram figure
#| code-fold: true
#| cache: true
#| fig-cap: "Weighted histograms of absenteeism outcomes."
print(g2_weighted_distribution)
```

```{r}
#| label: g21
#| code-summary: Zero-value profile comparison (frequency + weighted share)
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-2 zero-value profile by outcome: unweighted zero count and weighted zero proportion."
```

```{r}
#| label: g21-display
#| code-summary: Render zero-value profile comparison
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-2 zero-value profile by outcome: unweighted zero count and weighted zero proportion."
print(g21_zero_frequency_and_share)
```

```{r}
#| label: g22
#| code-summary: Core required statistic comparison across both outcomes
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-2 core required metrics by outcome: weighted mean/median/quartiles, weighted SD, and maximum value."
```

```{r}
#| label: g22-display
#| code-summary: Render core required statistic comparison
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-2 core required metrics by outcome: weighted mean/median/quartiles, weighted SD, and maximum value."
print(g22_required_core_comparison)
```

```{r}
#| label: g23
#| code-summary: Diagnostic required statistic comparison across both outcomes
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-2 diagnostic metrics by outcome: weighted variance, zero frequency, weighted zero proportion, and Pearson dispersion."
```

```{r}
#| label: g23-display
#| code-summary: Render diagnostic required statistic comparison
#| code-fold: true
#| cache: true
#| fig-cap: "Q4-2 diagnostic metrics by outcome: weighted variance, zero frequency, weighted zero proportion, and Pearson dispersion."
print(g23_required_diagnostic_comparison)
```

```{r}
#| label: verify-q42-proof
#| code-summary: "§4.2 PROOF: required metrics checklist and extracted verification table"
#| code-fold: true
q4_2_metrics_proof <- tibble::tibble(
  metric_required = c(
    "Mean (weighted)",
    "Standard error of mean (weighted)",
    "Median (weighted)",
    "Q1 (weighted)",
    "Q3 (weighted)",
    "Variance (weighted)",
    "Standard deviation (weighted)",
    "Frequency of zero values",
    "Proportion of zero values",
    "Maximum value",
    "Pearson dispersion (variance / mean)"
  ),
  computed_in_table = c(
    "weighted_mean" %in% names(q4_2_stats),
    "weighted_se_mean" %in% names(q4_2_stats),
    "weighted_median" %in% names(q4_2_stats),
    "weighted_q1" %in% names(q4_2_stats),
    "weighted_q3" %in% names(q4_2_stats),
    "weighted_variance" %in% names(q4_2_stats),
    "weighted_sd" %in% names(q4_2_stats),
    "zero_frequency_unweighted" %in% names(q4_2_stats),
    "zero_proportion_weighted" %in% names(q4_2_stats),
    "maximum_value" %in% names(q4_2_stats),
    "pearson_dispersion" %in% names(q4_2_stats)
  )
) %>%
  mutate(status = if_else(computed_in_table, "PASS ✓", "FAIL ⚠"))

verify_q42_proof <- q4_2_stats %>%
  select(
    outcome,
    weighted_mean,
    weighted_se_mean,
    weighted_q1,
    weighted_median,
    weighted_q3,
    weighted_variance,
    weighted_sd,
    zero_frequency_unweighted,
    zero_proportion_weighted,
    maximum_value,
    pearson_dispersion
  )

print(q4_2_metrics_proof)
print(verify_q42_proof)
```

::: {.callout-tip icon="false"}
**§4.2 DONE — Weighted descriptive profile completed** ✓

All requested weighted statistics were computed and displayed for both outcomes, with accompanying distribution and zero-value proportion visual checks.
:::

## B.4 Output Artifacts

| Output | Location | Content |
|------------------------|------------------------|------------------------|
| `q4_1_summary.parquet` | `data-private/derived/eda-3/` | Weighted comparison summary for Q4-1 |
| `q4_2_stats.parquet` | `data-private/derived/eda-3/` | Weighted descriptive statistics table for Q4-2 |
| Plot exports | `analysis/eda-3/prints/` | `g1`, `g11`, `g2`, `g21`, `g22`, `g23` high-resolution figures |

------------------------------------------------------------------------

# Session Info

```{r}
#| label: session-info
#| code-summary: Runtime and R session metadata
#| echo: false
#| code-fold: true
cat("Report completed at:", as.character(Sys.time()), "\n")
cat("Total runtime:", round(difftime(Sys.time(), report_render_start_time, units = "mins"), 2), "minutes\n")
sessionInfo()
```