---
 teh ---
title: "Chronic Work Absence Among Employed Canadians"
subtitle: "EDA 3 — CCHS 2010-11 & 2013-14: Distribution of Chronic Absence Days"
author: "Analysis Team"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This report explores how **chronic** work absence days (`absence_days_chronic`) are distributed among employed Canadian respondents in the CCHS 2010-11 and 2013-14 survey cycles. It replicates the EDA-2 analytical structure (same graph families, same demographic breakdowns) but targets the chronic-condition sub-measure rather than total absence.

-   **Dataset**: `cchs_employed` table from `cchs-3.sqlite` — each row is one employed survey respondent
-   **Research Question**: How many absence days attributable to a chronic health condition do employed Canadians report, and how does that distribution vary by sex, age, survey cycle, education, marital status, and immigration status?
-   **Data Lineage**: `ds0` → `ds1` (exclude NAs + 0-day chronic) → `chronic_ratio_tbl` (overview); `ds1` → `g1_data` → scatter + histogram; `ds1` → `g2_stats` / `sex_ratio_tbl` → faceted by sex; `ds1` → `g3_stats` → faceted by age group; `ds1` → `g4_data` / `g4_stats` → scatter + histogram by survey cycle; `ds1` → `ds5` (excl. NA education) → `g5_stats` → faceted by education level; `ds1` → `ds6` (excl. NA marital) → `g6_stats` → faceted by marital status; `ds1` → `ds7` (excl. NA immigration) → `g7_stats` → faceted by immigration status
-   **Scope**: Descriptive distribution of `absence_days_chronic`; no modelling in this EDA; directly comparable to EDA-2 (`absence_days_total`)

# Definition of Terms

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../') # keep this chunk isolated
```

```{r}
#| label: set_options
#| code-summary: report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
# set options shared by all chunks
opts_chunk$set(
  results    = 'show',
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = 'figure-png-iso/'
)
echo_chunks    <- FALSE # Toggle for debugging.
message_chunks <- FALSE # Toggle for debugging.
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-3/eda-3.R") # This allows knitr to call chunks from R script
```

# Environment

> Reviews the components of the working environment of the report. Non-technical readers are welcomed to skip.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

-   **ds0**: `cchs_employed` from `cchs-3.sqlite` — 64,248 employed respondents, one row each. Key column: `absence_days_chronic` (integer days missed due to a chronic health condition).
-   **ds1**: Filtered from `ds0` — respondents with 1+ chronic absence days only (0-day and NA respondents excluded). Working dataset for all analysis.
-   **g1_data**: Aggregated from `ds1` — one row per unique `absence_days_chronic` value with the count of respondents reporting that value.
-   **g2_stats**: Summarised from `ds1` — one row per `sex_label` with per-sex median, mean, and respondent count.
-   **sex_ratio_tbl**: Derived from `ds0` — per-sex count and percentage of respondents reporting zero vs. 1+ chronic absence days (among those who answered).
-   **g3_stats**: Summarised from `ds1` — one row per `age_group_3` with per-group median, mean, and respondent count. Saved to `data-local/g3_stats_age.csv`.
-   **g4_data**: Aggregated from `ds1` — one row per (`survey_cycle_label` × unique `absence_days_chronic` value) with respondent count; used for the g4 scatter.
-   **g4_stats**: Summarised from `ds1` — one row per `survey_cycle_label` with per-cycle median, mean, and respondent count.
-   **g5_stats**: Summarised from `ds5` (`ds1` minus NA `education_level`) — one row per `education_level` with per-group median, mean, and respondent count.
-   **g6_stats**: Summarised from `ds6` (`ds1` minus NA `marital_status_label`) — one row per `marital_status_label` with per-group median, mean, and respondent count.
-   **g7_stats**: Summarised from `ds7` (`ds1` minus NA `immigration_status_label`) — one row per `immigration_status_label` with per-group median, mean, and respondent count.

```{r}
#| label: load-data
#| code-summary: Load original data (ds0)
#| warning: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-0
#| code-summary: Initial data cleaning and coercion
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-1
#| code-summary: Create ds1 — exclude 0-day and NA chronic absence respondents
#| code-fold: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Data overview — ds0 structure
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-1
#| code-summary: Variable-level glimpse of ds0
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-2
#| code-summary: Key variable summary from ds0
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-3
#| code-summary: ds1 summary — central tendency of chronic absence days (1+ only)
#| code-fold: false
#| eval: true
#| cache: true
```

## Chronic Absence Overview: Who Reports Missing Days Due to a Chronic Condition?

Among employed Canadians who answered `absence_days_chronic`, many report **zero chronic absence days**. The table below shows the breakdown; the graphs in the Analysis section focus only on the `1+ days` subgroup (`ds1`).

```{r}
#| label: analytic-chronic-ratio
#| code-summary: Ratio of zero-chronic-absence vs. any-chronic-absence respondents
#| code-fold: true
#| eval: true
#| cache: true
```

# Analysis

## Graph Family g1: Distribution of Chronic Absence Days

### Data Preparation

`g1_data` aggregates `ds1` (respondents with 1+ chronic absence days) by unique values of `absence_days_chronic`, counting how many respondents reported each value. Median and mean are computed from `ds1` and appear as reference lines on both graphs below.

```{r}
#| label: g1-data-prep
#| code-summary: Prepare g1_data — frequency count per chronic absence-day value
#| code-fold: true
#| eval: true
#| cache: true
```

### Scatter Plot (g1-scatter)

Each point represents one unique value of `absence_days_chronic` (x-axis) plotted against the number of respondents with that value (y-axis). A dashed red line marks the median; a dotted orange line marks the mean. Both 0-day and NA respondents are excluded.

```{r}
#| label: g1-scatter
#| code-summary: Scatter — unique chronic day-values vs. respondent count
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Scatter of unique chronic absence-day values (x) vs. respondent count (y). X-axis zoomed to 1-40 via coord_cartesian (tail outside view). Breaks every 5 days. Dashed red = median, dotted orange = mean."
```

### Histogram (g1-hist)

Same population as the scatter (`ds1`, 1+ days only), binned into 5-day intervals and capped at 40 days for legibility. Median and mean reference lines are identical to the scatter.

```{r}
#| label: g1-hist
#| code-summary: Histogram — distribution of chronic absence days (5-day bins, zoomed to 1-40)
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Histogram of absence_days_chronic in 5-day bins, zoomed to 1-40 days via coord_cartesian (tail data not dropped). Breaks every 5 days. Dashed red = median, dotted orange = mean."
```

## Graph Family g2: Chronic Absence Days by Sex

### Data Preparation

`g2_stats` computes the median, mean, and respondent count for each `sex_label` group from `ds1`. These per-sex statistics drive the reference lines in the faceted histogram below.

```{r}
#| label: g2-data-prep
#| code-summary: Prepare g2_stats — per-sex central tendency from ds1
#| code-fold: true
#| eval: true
#| cache: true
```

### Chronic Absence Ratio by Sex

Among employed Canadians who answered `absence_days_chronic`, the table below shows what share of each sex group reported zero vs. one or more chronic absence days.

```{r}
#| label: analytic-sex-ratio
#| code-summary: Per-sex count and percentage of zero-chronic-absence vs. 1+ respondents
#| code-fold: true
#| eval: true
#| cache: true
```

### Histograms by Sex (g2-hist-sex)

Two panels side by side — one per sex. Each panel uses the same 5-day bins and 1-40 zoom as `g1-hist`, with its own median and mean reference lines.

```{r}
#| label: g2-hist-sex
#| code-summary: Faceted histogram — chronic absence day distribution split by sex_label
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Chronic absence day distribution by sex (5-day bins, zoomed 1-40). Dashed red = per-sex median, dotted orange = per-sex mean."
```

## Graph Family g3: Chronic Absence Days by Age Group

### Data Preparation

`g3_stats` computes the median, mean, and respondent count for each `age_group_3` from `ds1`. Parameters are also saved to `data-local/g3_stats_age.csv` for downstream use.

```{r}
#| label: g3-data-prep
#| code-summary: Prepare g3_stats — per-age-group central tendency from ds1
#| code-fold: true
#| eval: true
#| cache: true
```

### Histograms by Age Group (g3-hist-age)

One panel per `age_group_3`, laid out side by side. Same 5-day bins and 1-40 zoom as the prior families.

```{r}
#| label: g3-hist-age
#| code-summary: Faceted histogram — chronic absence day distribution by age_group_3
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Chronic absence day distribution by age group (5-day bins, zoomed 1-40). One panel per age_group_3; dashed red = per-group median, dotted orange = per-group mean."
```

## Graph Family g4: g1 Replicated by Survey Cycle

### Data Preparation

`g4_data` aggregates `ds1` within each `survey_cycle_label` to produce per-cycle frequency counts of `absence_days_chronic`. `g4_stats` provides per-cycle median and mean for the reference lines.

```{r}
#| label: g4-data-prep
#| code-summary: Prepare g4_data and g4_stats — per-cycle frequency and central tendency
#| code-fold: true
#| eval: true
#| cache: true
```

### Scatter Plot by Cycle (g4-scatter-cycle)

```{r}
#| label: g4-scatter-cycle
#| code-summary: Scatter by survey cycle — unique chronic day-values vs. respondent count per cycle
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Per-cycle scatter: each point is a unique chronic absence-day value within the cycle. Dashed red = per-cycle median, dotted orange = per-cycle mean. Zoomed to 1-40 days."
```

### Histogram by Cycle (g4-hist-cycle)

```{r}
#| label: g4-hist-cycle
#| code-summary: Histogram by survey cycle — 5-day bins, zoomed to 1-40
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Per-cycle histogram of absence_days_chronic (5-day bins, zoomed 1-40). Dashed red = per-cycle median, dotted orange = per-cycle mean."
```

## Graph Family g5: Chronic Absence Days by Education Level

### Data Preparation

`g5_stats` computes the median, mean, and respondent count for each `education_level` group from `ds5` — `ds1` with respondents missing `education_level` removed.

```{r}
#| label: g5-data-prep
#| code-summary: Prepare g5_stats — per-education-level central tendency from ds5
#| code-fold: true
#| eval: true
#| cache: true
```

### Histograms by Education Level (g5-hist-edu)

```{r}
#| label: g5-hist-edu
#| code-summary: Faceted histogram — chronic absence day distribution by education_level
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Chronic absence day distribution by education level (5-day bins, zoomed 1-40). One panel per education_level; dashed red = per-group median, dotted orange = per-group mean."
```

## Graph Family g6: Chronic Absence Days by Marital Status

### Data Preparation

`g6_stats` computes the median, mean, and respondent count for each `marital_status_label` group from `ds6` — `ds1` with respondents missing `marital_status_label` removed.

```{r}
#| label: g6-data-prep
#| code-summary: Prepare g6_stats — per-marital-status central tendency from ds6
#| code-fold: true
#| eval: true
#| cache: true
```

### Histograms by Marital Status (g6-hist-marital)

```{r}
#| label: g6-hist-marital
#| code-summary: Faceted histogram — chronic absence day distribution by marital_status_label
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Chronic absence day distribution by marital status (5-day bins, zoomed 1-40). One panel per marital_status_label; dashed red = per-group median, dotted orange = per-group mean."
```

## Graph Family g7: Chronic Absence Days by Immigration Status

### Data Preparation

`g7_stats` computes the median, mean, and respondent count for each `immigration_status_label` group from `ds7` — `ds1` with respondents missing `immigration_status_label` removed.

```{r}
#| label: g7-data-prep
#| code-summary: Prepare g7_stats — per-immigration-status central tendency from ds7
#| code-fold: true
#| eval: true
#| cache: true
```

### Histograms by Immigration Status (g7-hist-immigration)

```{r}
#| label: g7-hist-immigration
#| code-summary: Faceted histogram — chronic absence day distribution by immigration_status_label
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "Chronic absence day distribution by immigration status (5-day bins, zoomed 1-40). One panel per immigration_status_label; dashed red = per-group median, dotted orange = per-group mean."
```

# Session Info

```{r}
#| label: session-info
#| code-summary: R session information for reproducibility
#| echo: false
#| code-fold: true
cat("Report completed at:", as.character(Sys.time()), "\n")
cat("Total runtime:", round(difftime(Sys.time(), report_render_start_time, units = "mins"), 2), "minutes\n")
sessionInfo()
```

format: html: page-layout: full toc: true toc-location: right code-fold: show theme: yeti highlight-style: breeze code-line-numbers: true self-contained: true embed-resources: true keep-md: false editor: visual editor_options: chunk_output_type: console ---

# Mission

TODO: Describe the research question and analytical scope of this EDA.

-   **Dataset**: TODO
-   **Research Question**: TODO
-   **Data Lineage**: `ds0` → TODO
-   **Scope**: TODO

# Definition of Terms

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../') # keep this chunk isolated
```

```{r}
#| label: set_options
#| code-summary: report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
# set options shared by all chunks
opts_chunk$set(
  results    = 'show',
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = 'figure-png-iso/'
)
echo_chunks    <- FALSE # Toggle for debugging.
message_chunks <- FALSE # Toggle for debugging.
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-3/eda-3.R") # This allows knitr to call chunks from R script
```

# Environment

> Reviews the components of the working environment of the report. Non-technical readers are welcomed to skip.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

-   **ds0**: TODO — describe source table, unit of analysis, and key variables.
-   **ds1**: TODO — describe filtering/exclusion criteria applied to `ds0`.

```{r}
#| label: load-data
#| code-summary: Load original data (ds0)
#| warning: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-0
#| code-summary: Initial data cleaning and coercion
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-1
#| code-summary: Create ds1 — apply inclusion/exclusion criteria
#| code-fold: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Data overview — ds0 structure
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-1
#| code-summary: Variable-level glimpse of ds0
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-2
#| code-summary: Key variable summary from ds0
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-3
#| code-summary: ds1 summary — central tendency of key variable
#| code-fold: false
#| eval: true
#| cache: true
```

# Analysis

## Graph Family g1: TODO — describe focus

### Data Preparation

TODO: Describe what `g1_data` represents and how it is derived.

```{r}
#| label: g1-data-prep
#| code-summary: Prepare g1_data
#| code-fold: true
#| eval: true
#| cache: true
```

### Plot (g1-plot)

TODO: Describe the graph and what to look for.

```{r}
#| label: g1-plot
#| code-summary: Graph Family g1
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "TODO: add figure caption."
```

## Graph Family g2: TODO — describe focus

### Data Preparation

TODO: Describe what `g2_data` / `g2_stats` represents.

```{r}
#| label: g2-data-prep
#| code-summary: Prepare g2 data
#| code-fold: true
#| eval: true
#| cache: true
```

### Plot (g2-plot)

TODO: Describe the graph and what to look for.

```{r}
#| label: g2-plot
#| code-summary: Graph Family g2
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "TODO: add figure caption."
```

# Session Info

```{r}
#| label: session-info
#| code-summary: R session information for reproducibility
#| echo: false
#| code-fold: true
cat("Report completed at:", as.character(Sys.time()), "\n")
cat("Total runtime:", round(difftime(Sys.time(), report_render_start_time, units = "mins"), 2), "minutes\n")
sessionInfo()
```