---
title: "EDA 64 — Needs Variable Profile: Health Status & Functional Limitations"
subtitle: "Andersen Behavioral Model — Needs Domain"
author: "Analysis Team"
date: "2026-05-22"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: true
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This EDA profiles the 10 needs variables (perceived health status and functional
limitations) from the Andersen Behavioral Model framework. These represent the most
immediate determinant of health service use — the individual's perceived and evaluated
need for care.

**Research Questions:**

1. What is the weighted distribution of each needs variable in the employed population?
2. How does each needs variable relate to the primary outcome (`days_absent_total`) —
   mean days, median days, and zero-proportion?
3. What is the missing-data pattern across the 10 needs variables?
4. What health gradient patterns exist — perceived health × ADL functional limitation stacking?

ADL items are shown both individually (G1) and as a collapsed functional limitation
summary count (G2, G4).

Following the dual-file development pattern, all analytical code lives in
`analysis/eda-64/eda-64.R`. This document sources those chunks for presentation.

# Definition of Terms

**Needs factors**: In the Andersen Behavioral Model, these represent the most immediate
cause of health service use — perceived illness, evaluated health status, and functional
limitations (ADL).

**ADL (Activities of Daily Living)**: Six functional limitation domains: preparing meals,
running errands, housework, personal care, moving indoors, managing finances. Each coded
as ordinal (e.g., "NO", "SOMETIMES", "OFTEN").

**ADL limitation count**: Collapsed summary — number of ADL items (0–6) where the
respondent reports any limitation (value ≠ "NO").

**`days_absent_total`**: Primary outcome. Row-wise sum of all 8 LOP component variables,
representing total workdays missed in the past 3 months due to health reasons.

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = "../../")
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results   = "show",
  message   = FALSE,
  warning   = FALSE,
  comment   = NA,
  tidy      = FALSE,
  out.width = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-64/eda-64.R")
```

# Environment

> Reviews the components of the working environment. Non-technical readers may skip this section.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

```{r}
#| label: load-data
#| code-summary: Load data from Ellis parquet outputs
#| warning: false
#| code-fold: true
#| cache: true
```

## Data Context

For complete data documentation, see the
[Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: Tables and variables used in this analysis
#| code-fold: true
```

```{r}
#| label: data-context-person
#| code-summary: Representative individual with ADL limitations
#| code-fold: true
```

```{r}
#| label: data-context-distributions
#| code-summary: Level counts for needs variables
#| code-fold: true
```

```{r}
#| label: tweak-data-0
#| code-summary: Ensure factor types and create ADL summary
#| code-fold: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Data overview and ADL summary statistics
#| code-fold: true
```

# Analysis

## Graph Family G1: Variable Distributions

<!-- Research question: What is the weighted distribution of each needs variable? -->

```{r}
#| label: g1-data-prep
#| code-summary: Prepare distribution summaries
#| code-fold: true
#| cache: true
```

```{r}
#| label: g1
#| code-summary: Weighted distribution of health status variables
#| code-fold: true
#| fig-cap: "Population-weighted distribution of perceived health variables"
```

```{r}
#| label: g11
#| code-summary: Weighted prevalence of individual ADL limitations
#| code-fold: true
#| fig-cap: "Proportion reporting any limitation in each ADL domain"
```

```{r}
#| label: g12
#| code-summary: Weighted distribution by CCHS cycle
#| code-fold: true
#| fig-cap: "Health status distribution comparison between 2010-2011 and 2013-2014 cycles"
```

## Graph Family G2: Outcome Relationship

<!-- Research question: How does each needs variable relate to days_absent_total? -->

```{r}
#| label: g2-data-prep
#| code-summary: Prepare outcome summaries by variable level
#| code-fold: true
#| cache: true
```

```{r}
#| label: g2
#| code-summary: Mean days absent by health status variable level
#| code-fold: true
#| fig-cap: "Weighted mean days absent by perceived health level"
```

```{r}
#| label: g21
#| code-summary: Mean days absent by ADL limitation count (collapsed)
#| code-fold: true
#| fig-cap: "Dose-response: mean days absent increases with number of ADL limitations"
```

```{r}
#| label: g22
#| code-summary: Zero-day proportion by perceived health level
#| code-fold: true
#| fig-cap: "Proportion reporting zero days absent by self-rated health"
```

## Graph Family G3: Missing Data Profile

<!-- Research question: What is the missing-data pattern across needs variables? -->

```{r}
#| label: g3-data-prep
#| code-summary: Compute missingness statistics
#| code-fold: true
#| cache: true
```

```{r}
#| label: g3
#| code-summary: Missingness profile across needs variables
#| code-fold: true
#| fig-cap: "Percentage and count of missing values per variable"
```

```{r}
#| label: g31
#| code-summary: Missingness by CCHS cycle
#| code-fold: true
#| fig-cap: "Missing data rates by survey cycle"
```

## Graph Family G4: Health Gradient / ADL Stacking

<!-- Research question: What health gradient patterns exist? -->

```{r}
#| label: g4-data-prep
#| code-summary: Compute health × ADL cross-tabulation and co-occurrence
#| code-fold: true
#| cache: true
```

```{r}
#| label: g4
#| code-summary: Perceived health × ADL limitation count interaction
#| code-fold: true
#| fig-cap: "Mean days absent by perceived health and number of ADL limitations"
```

```{r}
#| label: g41
#| code-summary: ADL limitation co-occurrence heatmap
#| code-fold: true
#| fig-cap: "Which ADL limitation items tend to co-occur"
```
