---
title: "EDA 63 — Facilitating Variable Profile: Access & Health Behaviours"
subtitle: "Andersen Behavioral Model — Facilitating Domain"
author: "Analysis Team"
date: "2026-05-22"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: true
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This EDA profiles the 12 facilitating variables (health-system access and health
behaviours) from the Andersen Behavioral Model framework. These represent enabling
resources and behavioural choices that moderate the pathway between predisposition /
need and actual health service use (and work absence).

**Research Questions:**

1. What is the weighted distribution of each facilitating variable in the employed population?
2. How does each facilitating variable relate to the primary outcome (`days_absent_total`) —
   mean days, median days, and zero-proportion?
3. What is the missing-data pattern across the 12 facilitating variables?
4. What behaviour clustering patterns exist — smoking×BMI×activity co-profiles?

Following the dual-file development pattern, all analytical code lives in
`analysis/eda-63/eda-63.R`. This document sources those chunks for presentation.

# Definition of Terms

**Facilitating factors**: In the Andersen Behavioral Model, these are enabling resources
and conditions that permit or impede health service use — income, geography, insurance,
health care access — as well as health behaviours (smoking, physical activity, diet) that
moderate the disease–absence pathway.

**Province/Region**: Province is shown at both granular (13 levels) and regional grouping
(Atlantic, Quebec, Ontario, Prairies, British Columbia, Territories) levels for readability.

**`days_absent_total`**: Primary outcome. Row-wise sum of all 8 LOP component variables,
representing total workdays missed in the past 3 months due to health reasons.

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = "../../")
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results   = "show",
  message   = FALSE,
  warning   = FALSE,
  comment   = NA,
  tidy      = FALSE,
  out.width = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-63/eda-63.R")
```

# Environment

> Reviews the components of the working environment. Non-technical readers may skip this section.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

```{r}
#| label: load-data
#| code-summary: Load data from Ellis parquet outputs
#| warning: false
#| code-fold: true
#| cache: true
```

## Data Context

For complete data documentation, see the
[Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: Tables and variables used in this analysis
#| code-fold: true
```

```{r}
#| label: data-context-person
#| code-summary: Representative individual
#| code-fold: true
```

```{r}
#| label: data-context-distributions
#| code-summary: Level counts for facilitating variables
#| code-fold: true
```

```{r}
#| label: tweak-data-0
#| code-summary: Ensure factor types and add regional grouping
#| code-fold: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Data overview
#| code-fold: true
```

# Analysis

## Graph Family G1: Variable Distributions

<!-- Research question: What is the weighted distribution of each facilitating variable? -->

```{r}
#| label: g1-data-prep
#| code-summary: Prepare distribution summaries
#| code-fold: true
#| cache: true
```

```{r}
#| label: g1
#| code-summary: Unweighted distribution of facilitating variables
#| code-fold: true
#| fig-cap: "Unweighted frequency distribution by variable level"
#| fig-height: 8
```

```{r}
#| label: g11
#| code-summary: Weighted distribution — Province (13 levels)
#| code-fold: true
#| fig-cap: "Population-weighted distribution by province"
```

```{r}
#| label: g12
#| code-summary: Weighted distribution by CCHS cycle
#| code-fold: true
#| fig-cap: "Distribution comparison between 2010-2011 and 2013-2014 cycles"
#| fig-height: 8
```

## Graph Family G2: Outcome Relationship

<!-- Research question: How does each facilitating variable relate to days_absent_total? -->

```{r}
#| label: g2-data-prep
#| code-summary: Prepare outcome summaries by variable level
#| code-fold: true
#| cache: true
```

```{r}
#| label: g2
#| code-summary: Mean days absent by facilitating variable level
#| code-fold: true
#| fig-cap: "Weighted mean days absent by level"
#| fig-height: 8
```

```{r}
#| label: g21
#| code-summary: Mean days absent by province (granular, 13 levels)
#| code-fold: true
#| fig-cap: "Weighted mean days absent — all provinces sorted by value"
```

```{r}
#| label: g22
#| code-summary: Mean days absent by region (grouped)
#| code-fold: true
#| fig-cap: "Weighted mean days absent — provinces grouped into regions"
```

## Graph Family G3: Missing Data Profile

<!-- Research question: What is the missing-data pattern across facilitating variables? -->

```{r}
#| label: g3-data-prep
#| code-summary: Compute missingness statistics
#| code-fold: true
#| cache: true
```

```{r}
#| label: g3
#| code-summary: Missingness profile across facilitating variables
#| code-fold: true
#| fig-cap: "Percentage and count of missing values per variable"
```

```{r}
#| label: g31
#| code-summary: Missingness by CCHS cycle
#| code-fold: true
#| fig-cap: "Missing data rates by survey cycle"
```

## Graph Family G4: Behaviour Clustering

<!-- Research question: What behaviour clustering patterns exist? -->

```{r}
#| label: g4-data-prep
#| code-summary: Compute behaviour cross-tabulation summaries
#| code-fold: true
#| cache: true
```

```{r}
#| label: g4
#| code-summary: Smoking status × BMI category interaction
#| code-fold: true
#| fig-cap: "Mean days absent by smoking status and BMI"
```

```{r}
#| label: g41
#| code-summary: Smoking status × physical activity interaction
#| code-fold: true
#| fig-cap: "Mean days absent by smoking status and physical activity level"
```
