---
title: "Work Absence Among Employed Canadians"
subtitle: "EDA 2 — CCHS 2010-11 & 2013-14: Days Missed by Employed Respondents"
author: "Analysis Team"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mission

> **TODO**: Replace this section with the analytical mission for EDA-2.

Describe:

-   **Dataset**: What data are you analysing and where does it come from?
-   **Research Questions**: What are you trying to learn?
-   **Data Lineage**: What derived datasets will be created (e.g. `ds0 → g2_data → ...`)?
-   **Scope**: What is in and out of scope for this EDA?

# Definition of Terms

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false 
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../') # keep this chunk isolated
```

```{r}
#| label: set_options
#| code-summary: report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
# set options shared by all chunks
opts_chunk$set(
  results    = 'show',
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = 'figure-png-iso/'
)
echo_chunks    <- FALSE # Toggle for debugging.
message_chunks <- FALSE # Toggle for debugging.
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-2/eda-2.R") # This allows knitr to call chunks from R script
```

# Environment

> Reviews the components of the working environment of the report. Non-technical readers are welcomed to skip.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

> **TODO**: Describe the data lineage for EDA-2.

-   **ds0**: Original source data — *replace with your description*
-   **g2_data**: Derived summary created in `g2-data-prep` chunk — *replace with your description*

```{r}
#| label: load-data
#| code-summary: Load original data (ds0)
#| warning: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: tweak-data-0
#| code-summary: Initial data cleaning and coercion
#| code-fold: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Data overview — ds0 structure
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-1
#| code-summary: Variable-level glimpse of ds0
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: inspect-data-2
#| code-summary: Key variable summary from ds0
#| code-fold: false
#| eval: true
#| cache: true
```

# Analysis

## Exploratory Visualization

### First Look at the Data (g1)

> **TODO**: Describe what this graph shows and why it is the first thing you look at.

```{r}
#| label: g1
#| code-summary: Initial exploratory graph using original data (ds0)
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "TODO: Replace with a descriptive caption for g1"
```

## Graph Family g2x: \[Replace with Family Theme\]

### Data Preparation

> **TODO**: Describe what analytical question the g2 family explores and what the shared data ancestor represents.

```{r}
#| label: g2-data-prep
#| code-summary: Prepare g2_data — shared ancestor for all g2x visualizations
#| code-fold: true
#| eval: true
#| cache: true
```

### Primary View (g2)

> **TODO**: Describe what this graph shows.

```{r}
#| label: g2
#| code-summary: Primary g2 family visualization
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "TODO: Replace with a descriptive caption for g2"
```

### Alternative Perspective (g21)

> **TODO**: Describe how this view differs from g2 while using the same data ancestor.

```{r}
#| label: g21
#| code-summary: Second g2 family member — same data, different analytical lens
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "TODO: Replace with a descriptive caption for g21"
```

````{=html}
<!-- 
Add more graph family members here following the same pattern:

### Another Family Member (g22)

```{r}
#| label: g22
#| code-summary: ...
#| code-fold: true
#| eval: true
#| cache: true
#| fig-cap: "..."
```

Add new families (g3, g4, ...) by repeating the
  data-prep + family-member chunk structure above.
-->
````

# Session Info

```{r}
#| label: session-info
#| code-summary: R session information for reproducibility
#| echo: false
#| code-fold: true
cat("Report completed at:", as.character(Sys.time()), "\n")
cat("Total runtime:", round(difftime(Sys.time(), report_render_start_time, units = "mins"), 2), "minutes\n")
sessionInfo()
```