---
title: "Pipeline Observation: Ferry & Ellis Transformation Report"
subtitle: "EDA 2 — CCHS Data Pipeline: 1-ferry.R and 2-ellis.R"
author: "Andriy Koval"
date: "2026-03-05"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This document **observes and documents** the two data manipulation scripts that build the CCHS analytical dataset used in the study *"Predictors of Work Absenteeism Associated with Chronic Conditions Among Canadian Workers"*. It does not perform statistical analysis — it tracks what each script does, how many rows and variables are affected at each step, and ties every operation back to the requirements in `stats_instructions_v3.md`.

**Data genealogy:**

```         
CCHS2010_LOP.sav  ──┐
                     ├──► 1-ferry.R ──► cchs-1.sqlite ──► 2-ellis.R ──► cchs_analytical.parquet
CCHS_2014_EN_PUMF.sav ──┘                                              └──► sample_flow.parquet
```

This report is split into two parts:

-   **Part A — Ferry Observation**: What `1-ferry.R` receives, what it does, and what it produces
-   **Part B — Ellis Observation**: What `2-ellis.R` receives, all transformations applied, and what the final analytical dataset contains — including the effect of different configuration flag scenarios

# Environment

```{r}
#| label: knitr-root
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../')
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = 'show',
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = 'figure-png-iso/'
)
echo_chunks    <- FALSE
message_chunks <- FALSE
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-2/eda-2.R")
```

```{r}
#| label: load-packages
#| code-summary: Packages used in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Project-level functions
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Path constants and white-list counts
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Helper functions (with_data, etc.)
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

------------------------------------------------------------------------

# Part A — Ferry Observation (`1-ferry.R`) {#part-a}

## What is the Ferry Pattern?

The **Ferry Pattern** transports raw data from its source format into a local staging store with **zero semantic transformation**. The ferry script is intentionally limited: it is forbidden from renaming columns (beyond `clean_names()` sanitization), selecting subsets, recoding values, or excluding rows. Every such decision is deferred to the Ellis script.

::: {.callout-note icon="false"}
**Requirement — §2.1 Database** (`stats_instructions_v3.md`):

> The analysis uses two cycles of the Canadian Community Health Survey (CCHS), a publicly available, anonymized, nationally representative survey conducted by Statistics Canada:
>
> -   **CCHS 2010-2011** (Annual Component)
> -   **CCHS 2013-2014** (Annual Component)
>
> The relevant modules required are: **Chronic Conditions module (CCC)**, **Loss of Productivity module (LOP)**, and **demographic and socioeconomic variables** from the main questionnaire.
:::

## A.1 Inputs

| File                    | Description                          | Format      |
|------------------------|------------------------|------------------------|
| `CCHS2010_LOP.sav`      | CCHS 2010-2011 Annual Component PUMF | SPSS (.sav) |
| `CCHS_2014_EN_PUMF.sav` | CCHS 2013-2014 Annual Component PUMF | SPSS (.sav) |

Both files reside in `data-private/raw/2026-02-19/` and are not version-controlled (private microdata).

## A.2 Operations

Three lightweight technical operations are applied to each file — no data values are changed:

| Step | Function | What it does |
|------------------------|------------------------|------------------------|
| 1 | `haven::read_sav()` | Reads SPSS binary into an R data frame, preserving all columns and numeric codes |
| 2 | `haven::zap_labels()` + `haven::zap_label()` | Strips SPSS value labels and variable labels — SQLite cannot store them; numeric codes are retained |
| 3 | `janitor::clean_names()` | Lowercases all column names and replaces spaces/special characters with underscores (e.g., `LOPG040` → `lopg040`) |

No rows are removed. No values are changed. No columns are added or dropped.

## A.3 Output Size

```{r}
#| label: ferry-load-data
#| code-summary: Load ferry metadata from cchs-1.sqlite
#| code-fold: true
#| cache: false
```

```{r}
#| label: g-ferry-size
#| code-summary: "Chart: rows and columns per cycle table"
#| code-fold: true
#| fig-cap: "Ferry output size — each cycle table as written to cchs-1.sqlite. The CCHS PUMF files contain hundreds of variables; all are retained by the ferry script with no selection."
```

## A.4 Column Inventory

```{r}
#| label: inspect-ferry-columns
#| code-summary: Column distribution by module prefix + overlap stats
#| code-fold: false
#| cache: false
```

The table below shows how many columns fall under each CCHS module prefix (e.g., `ccc_*`, `lop*`, `dhh*`). The two cycles share most columns but differ on a small number of cycle-specific derived variables.

## A.5 Data Preview

::: {.callout-important icon="false"}
**Note on scale**: The full CCHS PUMF contains 600+ columns per cycle. The preview below shows only the **13 confirmed required variables** from `required-variables-and-sample.md` to keep the output manageable. The ferry output retains all columns.
:::

```{r}
#| label: inspect-ferry-sample
#| code-summary: First 5 rows — confirmed columns only
#| code-fold: false
#| cache: false
```

## A.6 Ferry Output Summary

| Output | Location | Content |
|------------------------|------------------------|------------------------|
| `cchs-1.sqlite` | `data-private/derived/` | Primary staging — tables `cchs_2010_raw` + `cchs_2014_raw` |
| `cchs_2010_raw.parquet` | `data-private/derived/cchs-1-raw/` | Parquet backup of 2010-2011 cycle |
| `cchs_2014_raw.parquet` | `data-private/derived/cchs-1-raw/` | Parquet backup of 2013-2014 cycle |

------------------------------------------------------------------------

# Part B — Ellis Observation (`2-ellis.R`) {#part-b}

## What is the Ellis Pattern?

The **Ellis Pattern** transforms the ferry's staging output into an analysis-ready dataset. It applies white-list variable selection, cross-cycle harmonization, factor recoding, and sample exclusions — exactly the operations the ferry was forbidden from doing.

## B.1 Inputs & Configuration Flags

**Input**: `data-private/derived/cchs-1.sqlite`\
Tables: `cchs_2010_raw`, `cchs_2014_raw` (all columns; numeric codes)

Three boolean flags at the top of `2-ellis.R` control which transformations are applied:

| Flag | Default Scenario | All-TRUE Scenario | Effect |
|------------------|------------------|------------------|------------------|
| `strict_cycle_integrity` | `FALSE` | `TRUE` | If TRUE, stops on error when either cycle table is empty; if FALSE, continues with a warning |
| `apply_sample_exclusions` | `TRUE` | `TRUE` | Apply §3.1 age / employment / proxy / outcome-completeness exclusions |
| `apply_completeness_exclusion` | `FALSE` | `TRUE` | Also exclude respondents with any missing value on CCC conditions or predictor variables (full §3.1 criterion 4b) |

This report documents the **Default Scenario** with real output and describes the **All-TRUE Scenario** analytically.

## B.2 White-List Variable Selection

::: {.callout-note icon="false"}
**Requirement — §2.2 Required Variables** (`stats_instructions_v3.md`):

> The following variables (or their CCHS equivalents across cycles) must be identified and extracted. Variable names may differ slightly between cycles; consult the CCHS data dictionaries.

| Category | Variables |
|------------------------------------|------------------------------------|
| Outcome | Days absent (LOP module sum); sensitivity outcome (chronic condition days only) |
| Primary exposure | 19 binary chronic conditions (CCC module) |
| Predisposing | Age, sex, marital status, household size, children counts, education, immigration, ethnicity, homeownership, student status |
| Facilitating | Income, province, family doctor, employment type, work schedule, occupation, job stress, diet, alcohol, smoking, BMI, physical activity |
| Needs | Functional limitations, injury, self-perceived health (general + mental + change) |
| Survey design | Master weight, 500 bootstrap replicates, strata/cluster IDs |
:::

The white-list approach selects \~69 named variables plus \~500 bootstrap weight columns from the 600+ column ferry output. Variables are classified as:

-   **Confirmed** (13): explicitly verified in `data-public/derived/required-variables-and-sample.md`
-   **Inferred — CCC** (19): derived from CCHS PUMF naming conventions; trigger a WARNING if absent
-   **Inferred — other** (\~37): predisposing, facilitating, needs, and ID variables; also trigger a WARNING if absent

```{r}
#| label: g-whitelist-ratio
#| code-summary: "Chart: white-list selection ratio"
#| code-fold: true
#| fig-cap: "Column selection ratio per cycle. The confirmed and inferred variables account for roughly 10-12% of the total columns in each SPSS file; the remaining ~88% are not selected by 2-ellis.R."
```

## B.3 Alias Harmonization

Before white-list selection, `2-ellis.R` maps known cross-cycle variable name aliases to canonical names. If the canonical name is missing but an alias is found, the alias is copied in:

| Canonical (white-list name) | Aliases checked      |
|-----------------------------|----------------------|
| `edudh04`                   | `edudr04`            |
| `sdcdgcb`                   | `sdcgcgt`            |
| `geodgprv`                  | `geogprv`            |
| `hcu_1aa`                   | `hcu_1a`, `hcudgmd`  |
| `lbfdghp`                   | `lbsg031`, `lbs_g31` |
| `gen_02a`                   | `gen_02`, `gen_02a2` |
| `gen_09`                    | `gen_03`             |
| `inj_01`                    | `injdgyrs`           |

## B.4 Step 1 — Outcome Construction

::: {.callout-note icon="false"}
**Requirement — §4.1 Construction** (`stats_instructions_v3.md`):

> The primary outcome is the total number of workdays missed in the three months preceding the survey due to any health-related reason. This is a count variable derived from the LOP module by summing days absent across all reported health reasons: *Chronic condition, Injury, Cold, Gastroenteritis, Flu/influenza, Other respiratory infection, Other infectious disease, Any other reason*.
>
> The previous analysis specifies a minimum value of 1 day and a maximum of 90 days.
:::

Two outcome variables are constructed:

| Variable | Logic | Note |
|------------------------|------------------------|------------------------|
| `days_absent_total` | `rowSums` of 8 LOP components (`lopg040` + `lopg070` + `lopg082`–`lopg086` + `lopg100`) | NA-safe: at least one non-NA component must exist; all-NA → treated as structural 0 |
| `days_absent_chronic` | `as.numeric(lopg040)` | Sensitivity outcome — chronic condition days only |

Values outside the 0–90 range are set to `NA` and flagged as data quality issues.

## B.5 Step 2 — Sample Exclusions

::: {.callout-note icon="false"}
**Requirement — §3.1 Inclusion and Exclusion Criteria** (`stats_instructions_v3.md`):

> Apply the following exclusion criteria in order, recording the sample size at each step:
>
> 1.  Exclude respondents aged \< 15 years or \> 75 years
> 2.  Exclude respondents who did not report paid employment in the past three months
> 3.  Exclude proxy respondents (i.e., where someone else answered on the respondent's behalf)
> 4.  Exclude respondents with incomplete answers on the outcome variable (LOP total days), any of the 19 chronic conditions, or any of the predictor variables
>
> The previous analysis reports a **final unweighted sample of n = 64,141** from an initial pool of 62,909 (CCHS 2011) + 63,522 (CCHS 2014).
:::

### Default Scenario (`apply_sample_exclusions = TRUE`, `apply_completeness_exclusion = FALSE`)

Steps 1–3 plus outcome completeness are applied. Step 4b (CCC + predictor completeness) is deferred to the analysis phase.

```{r}
#| label: ellis-load-data
#| code-summary: Load analytical parquet and sample_flow
#| code-fold: true
#| cache: false
```

```{r}
#| label: g-exclusion-funnel
#| code-summary: "Chart: sample exclusion funnel"
#| code-fold: true
#| fig-cap: "Sample exclusion funnel — n remaining after each sequential filter. The employment filter (step 3) is the largest single reduction, reflecting that the CCHS surveys the general population while the study focuses only on employed respondents."
```

### All-TRUE Scenario (`apply_completeness_exclusion = TRUE`)

When `apply_completeness_exclusion = TRUE`, an additional step is inserted after complete-outcome filtering:

> **Step 4b**: Exclude respondents with any `NA` on the 19 CCC conditions or on any predictor variable (full §3.1 criterion 4b compliance).

This brings the final sample closer to the thesis target of **n = 64,141**. The default scenario retains a larger sample and handles missingness at the modelling stage.

| Scenario | Estimated final n | Completeness criterion |
|------------------------|------------------------|------------------------|
| Default (`apply_completeness_exclusion = FALSE`) | \~70,000 | Outcome only |
| All-TRUE (`apply_completeness_exclusion = TRUE`) | \~64,141 (§3.1 target) | Outcome + all CCC + all predictors |

## B.6 Step 3 — Cycle Pooling & Harmonization

::: {.callout-note icon="false"}
**Requirement — §3.2 Pooling the Two Cycles** (`stats_instructions_v3.md`):

> -   Harmonize variable names and coding schemes between the two cycles (they may differ)
> -   Create a cycle indicator variable (0 = 2010-2011, 1 = 2013-2014)
> -   **Adjust survey weights**: when combining two cycles, divide each respondent's original survey weight by 2. This is the standard Statistics Canada recommendation for pooling CCHS annual cycles.
> -   Verify missing values that might be logically imputed. Some questions are only asked of respondents aged 18 years and older.
:::

Steps applied:

1.  `cycle = 0L` added to CCHS 2010-2011 rows; `cycle = 1L` to CCHS 2013-2014
2.  Column union: any column present in only one cycle is added as `NA` in the other
3.  `bind_rows()` stacks both cycles into a single pooled data frame `ds0`

```{r}
#| label: g-cycle-split
#| code-summary: "Chart: pooling balance between cycles"
#| code-fold: true
#| fig-cap: "Proportion of the final analytical sample contributed by each CCHS cycle. Statistics Canada's recommendation is to weight each cycle's contribution equally (÷ 2 applied to survey weights) when pooling two annual cycles."
```

## B.7 Step 4 — Factor Recoding

All raw CCHS numeric codes are recoded into labelled R factors. CCHS special codes (6 = Not applicable, 7 = Don't know, 8 = Refusal, 9 = Not stated) are mapped to `NA`. The resulting `ds3` object contains human-readable factor levels in place of integer codes.

Key recoded variables:

| Raw variable | Recoded variable | Levels |
|------------------------|------------------------|------------------------|
| `dhhgage` | `age_group` | `15-24` / `25-54` / `55-75` (ordered) |
| `dhh_sex` | `sex` | `Male` / `Female` |
| `dhhgms` | `marital_status` | Single / Married / Common-law / Widowed-Divorced-Separated |
| `edudh04` | `education` | 4 ordered levels (\< secondary → post-secondary graduate) |
| `incdghh` | `income_5cat` | 5 ordered income bands (\< \$20k → \$80k+) |
| `smkdsty` | `smoking_status` | Never / Former / Occasional / Daily (ordered) |
| `hwtdgbmi` | `bmi_category` | Underweight / Normal / Overweight / Obese (ordered) |
| `ccc_015`–`ccc_300` | `ccc_015_yn`–`ccc_300_yn` | `Yes` / `No` (all 19 CCC conditions) |

```{r}
#| label: inspect-analytical
#| code-summary: Analytical dataset structure and cycle distribution
#| code-fold: false
#| cache: false
```

## B.8 Outcome Distribution

::: {.callout-note icon="false"}
**Requirement — §4.2 Distribution Checks** (`stats_instructions_v3.md`):

> Before modeling, verify and report: mean and SE (weighted), median/Q1/Q3 (weighted), frequency and proportion of zero values, maximum value, Pearson dispersion statistic.
>
> The previous analysis reports: **mean = 1.35 (SE = 0.02)**, **70.59% zeros**, variance = 17.7, SD = 4.19.
:::

```{r}
#| label: g2-data-prep
#| code-summary: Bin days_absent_total into categories for display
#| code-fold: true
#| cache: false
```

```{r}
#| label: g2
#| code-summary: "Chart: outcome distribution"
#| code-fold: true
#| fig-cap: "Distribution of days absent from work (binned). The outcome is strongly zero-inflated — approximately 70% of employed respondents reported no health-related absences in the past 3 months, motivating a zero-inflated or negative-binomial modelling approach."
```

## B.9 Ellis Output Summary

| Output | Location | Format | Content |
|------------------|------------------|------------------|------------------|
| `cchs_analytical.parquet` | `data-private/derived/cchs-2-tables/` | Parquet | Pooled analytical dataset with recoded factors |
| `sample_flow.parquet` | `data-private/derived/cchs-2-tables/` | Parquet | Exclusion step counts for Figure 1 flowchart |
| `cchs_analytical` table | `data-private/derived/cchs-2.sqlite` | SQLite | Same data, factors as character (for ad-hoc SQL) |
| `sample_flow` table | `data-private/derived/cchs-2.sqlite` | SQLite | Same flowchart data |

------------------------------------------------------------------------

# Report Metadata

```{r}
#| echo: false
cat("Report rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Elapsed:", round(difftime(Sys.time(), report_render_start_time, units = "secs"), 1), "seconds\n")
```