---
title: "Pipeline Observation: Ferry & Ellis Transformation Report"
subtitle: "EDA 2 — CCHS Data Pipeline: 1-ferry.R and 2-ellis.R"
author: "Oleksandr Koval"
date: "2026-03-05"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This document **observes and documents** the two data manipulation scripts that build the CCHS analytical dataset used in the study *"Predictors of Work Absenteeism Associated with Chronic Conditions Among Canadian Workers"*. It does not perform statistical analysis — it tracks what each script does, how many rows and variables are affected at each step, and ties every operation back to the requirements in `stats_instructions_v3.md`.

**Data genealogy:**

```         
CCHS2010_LOP.sav  ──┐
                     ├──► 1-ferry.R ──► cchs-1.sqlite ──► 2-ellis.R ──► cchs_analytical.parquet
CCHS_2014_EN_PUMF.sav ──┘                                              └──► sample_flow.parquet
```

This report is split into two parts:

-   **Part A — Ferry Observation**: What `1-ferry.R` receives, what it does, and what it produces
-   **Part B — Ellis Observation**: What `2-ellis.R` receives, all transformations applied, and what the final analytical dataset contains — including the effect of different configuration flag scenarios

# Environment

```{r}
#| label: knitr-root
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../')
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = 'show',
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = 'figure-png-iso/'
)
echo_chunks    <- FALSE
message_chunks <- FALSE
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-2/eda-2.R")
```

```{r}
#| label: load-packages
#| code-summary: Packages used in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Project-level functions
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Path constants and white-list counts
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Helper functions (with_data, etc.)
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

------------------------------------------------------------------------

# Part A — Ferry Observation (`1-ferry.R`) {#part-a}

## What is the Ferry Pattern?

The **Ferry Pattern** transports raw data from its source format into a local staging store with **zero semantic transformation**. The ferry script is intentionally limited: it is forbidden from renaming columns (beyond `clean_names()` sanitization), selecting subsets, recoding values, or excluding rows. Every such decision is deferred to the Ellis script.

::: {.callout-note icon="false"}
**Requirement — §2.1 Database** (`stats_instructions_v3.md`):

> The analysis uses two cycles of the Canadian Community Health Survey (CCHS), a publicly available, anonymized, nationally representative survey conducted by Statistics Canada:
>
> -   **CCHS 2010-2011** (Annual Component)
> -   **CCHS 2013-2014** (Annual Component)
>
> The relevant modules required are: **Chronic Conditions module (CCC)**, **Loss of Productivity module (LOP)**, and **demographic and socioeconomic variables** from the main questionnaire.
:::

## A.1 Inputs

| File                    | Description                          | Format      |
|------------------------|------------------------|------------------------|
| `CCHS2010_LOP.sav`      | CCHS 2010-2011 Annual Component PUMF | SPSS (.sav) |
| `CCHS_2014_EN_PUMF.sav` | CCHS 2013-2014 Annual Component PUMF | SPSS (.sav) |

Both files reside in `data-private/raw/2026-02-19/` and are not version-controlled (private microdata).

## A.2 Operations

Three lightweight technical operations are applied to each file — no data values are changed:

| Step | Function | What it does |
|------------------------|------------------------|------------------------|
| 1 | `haven::read_sav()` | Reads SPSS binary into an R data frame, preserving all columns and numeric codes |
| 2 | `haven::zap_labels()` + `haven::zap_label()` | Strips SPSS value labels and variable labels — SQLite cannot store them; numeric codes are retained |
| 3 | `janitor::clean_names()` | Lowercases all column names and replaces spaces/special characters with underscores (e.g., `LOPG040` → `lopg040`) |

No rows are removed. No values are changed. No columns are added or dropped.

## A.3 Output Size

```{r}
#| label: ferry-load-data
#| code-summary: Load ferry metadata from cchs-1.sqlite
#| code-fold: true
#| cache: false
```

```{r}
#| label: g-ferry-size
#| code-summary: "Chart: rows and columns per cycle table"
#| code-fold: true
#| fig-cap: "Ferry output size — each cycle table as written to cchs-1.sqlite. The CCHS PUMF files contain hundreds of variables; all are retained by the ferry script with no selection."
```

```{r}
#| label: verify-ferry-cycles
#| code-summary: "\u00a72.1 PROOF: both CCHS cycles loaded with expected row counts"
#| code-fold: true
```

::: {.callout-tip icon="false"}
**§2.1 DONE — Both CCHS cycles loaded** ✓

The ferry output `cchs-1.sqlite` contains two tables. The starting pool matches §3.1: `cchs_2010_raw` = 62,909 rows and `cchs_2014_raw` = 63,522 rows, giving a combined pool of **126,431 respondents** before any exclusions.
:::

## A.4 Column Inventory

```{r}
#| label: inspect-ferry-columns
#| code-summary: Column distribution by module prefix + overlap stats
#| code-fold: false
#| cache: false
```

The table below shows how many columns fall under each CCHS module prefix (e.g., `ccc_*`, `lop*`, `dhh*`). The two cycles share most columns but differ on a small number of cycle-specific derived variables.

```{r}
#| label: verify-modules-present
#| code-summary: "\u00a72.1 PROOF: CCC, LOP, and demographic modules present in both cycles"
#| code-fold: true
```

::: {.callout-tip icon="false"}
**§2.1 DONE — All required CCHS modules present** ✓

Column prefix scan confirms that `ccc_*` (Chronic Conditions), `lop*` (Loss of Productivity), `dhh*` (demographics), `bsw*` (bootstrap weights), and `geo*` (province) groups are all present in both CCHS cycles inside the ferry output — satisfying the §2.1 module requirements.
:::

## A.5 Data Preview

::: {.callout-important icon="false"}
**Note on scale**: The full CCHS PUMF contains 600+ columns per cycle. The preview below shows only the **13 confirmed required variables** from `required-variables-and-sample.md` to keep the output manageable. The ferry output retains all columns.
:::

```{r}
#| label: inspect-ferry-sample
#| code-summary: First 5 rows — confirmed columns only
#| code-fold: false
#| cache: false
```

## A.6 Ferry Output Summary

| Output | Location | Content |
|------------------------|------------------------|------------------------|
| `cchs-1.sqlite` | `data-private/derived/` | Primary staging — tables `cchs_2010_raw` + `cchs_2014_raw` |
| `cchs_2010_raw.parquet` | `data-private/derived/cchs-1-raw/` | Parquet backup of 2010-2011 cycle |
| `cchs_2014_raw.parquet` | `data-private/derived/cchs-1-raw/` | Parquet backup of 2013-2014 cycle |

------------------------------------------------------------------------

# Part B — Ellis Observation (`2-ellis.R`) {#part-b}

## What is the Ellis Pattern?

The **Ellis Pattern** transforms the ferry's staging output into an analysis-ready dataset. It applies white-list variable selection, cross-cycle harmonization, factor recoding, and sample exclusions — exactly the operations the ferry was forbidden from doing.

## B.1 Inputs & Configuration Flags

**Input**: `data-private/derived/cchs-1.sqlite`\
Tables: `cchs_2010_raw`, `cchs_2014_raw` (all columns; numeric codes)

Three boolean flags at the top of `2-ellis.R` control which transformations are applied:

| Flag | Default Scenario | All-TRUE Scenario | Effect |
|------------------|------------------|------------------|------------------|
| `strict_cycle_integrity` | `FALSE` | `TRUE` | If TRUE, stops on error when either cycle table is empty; if FALSE, continues with a warning |
| `apply_sample_exclusions` | `TRUE` | `TRUE` | Apply §3.1 age / employment / proxy / outcome-completeness exclusions |
| `apply_completeness_exclusion` | `FALSE` | `TRUE` | Also exclude respondents with any missing value on CCC conditions or predictor variables (full §3.1 criterion 4b) |

This report documents the **Default Scenario** with real output and describes the **All-TRUE Scenario** analytically.

## B.2 White-List Variable Selection

::: {.callout-note icon="false"}
**Requirement — §2.2 Required Variables** (`stats_instructions_v3.md`):

> The following variables (or their CCHS equivalents across cycles) must be identified and extracted. Variable names may differ slightly between cycles; consult the CCHS data dictionaries.

| Category | Variables |
|------------------------------------|------------------------------------|
| Outcome | Days absent (LOP module sum); sensitivity outcome (chronic condition days only) |
| Primary exposure | 19 binary chronic conditions (CCC module) |
| Predisposing | Age, sex, marital status, household size, children counts, education, immigration, ethnicity, homeownership, student status |
| Facilitating | Income, province, family doctor, employment type, work schedule, occupation, job stress, diet, alcohol, smoking, BMI, physical activity |
| Needs | Functional limitations, injury, self-perceived health (general + mental + change) |
| Survey design | Master weight, 500 bootstrap replicates, strata/cluster IDs |
:::

The white-list approach selects \~69 named variables plus \~500 bootstrap weight columns from the 600+ column ferry output. Variables are classified as:

-   **Confirmed** (13): explicitly verified in `data-public/derived/required-variables-and-sample.md`
-   **Inferred — CCC** (19): derived from CCHS PUMF naming conventions; trigger a WARNING if absent
-   **Inferred — other** (\~37): predisposing, facilitating, needs, and ID variables; also trigger a WARNING if absent

```{r}
#| label: g-whitelist-ratio
#| code-summary: "Chart: white-list selection ratio"
#| code-fold: true
#| fig-cap: "Column selection ratio per cycle. The confirmed and inferred variables account for roughly 10-12% of the total columns in each SPSS file; the remaining ~88% are not selected by 2-ellis.R."
```

## B.3 Alias Harmonization

Before white-list selection, `2-ellis.R` maps known cross-cycle variable name aliases to canonical names. If the canonical name is missing but an alias is found, the alias is copied in:

| Canonical (white-list name) | Aliases checked      |
|-----------------------------|----------------------|
| `edudh04`                   | `edudr04`            |
| `sdcdgcb`                   | `sdcgcgt`            |
| `geodgprv`                  | `geogprv`            |
| `hcu_1aa`                   | `hcu_1a`, `hcudgmd`  |
| `lbfdghp`                   | `lbsg031`, `lbs_g31` |
| `gen_02a`                   | `gen_02`, `gen_02a2` |
| `gen_09`                    | `gen_03`             |
| `inj_01`                    | `injdgyrs`           |

## B.4 Step 2 — Sample Exclusions

```{r}
#| label: ellis-load-data
#| code-summary: Load analytical parquet and sample_flow
#| code-fold: true
#| cache: false
#| include: false
```


::: {.callout-note icon="false"}
**Requirement — §3.1 Inclusion and Exclusion Criteria** (`stats_instructions_v3.md`):

> Apply the following exclusion criteria in order, recording the sample size at each step:
>
> 1.  Exclude respondents aged \< 15 years or \> 75 years
> 2.  Exclude respondents who did not report paid employment in the past three months
> 3.  Exclude proxy respondents (i.e., where someone else answered on the respondent's behalf)
> 4.  Exclude respondents with incomplete answers on the outcome variable (LOP total days), any of the 19 chronic conditions, or any of the predictor variables
>
> The previous analysis reports a **final unweighted sample of n = 64,141** from an initial pool of 62,909 (CCHS 2011) + 63,522 (CCHS 2014).
:::

### Default Scenario (`apply_sample_exclusions = TRUE`, `apply_completeness_exclusion = FALSE`)

Steps 1–3 plus outcome completeness are applied. Step 4b (CCC + predictor completeness) is deferred to the analysis phase.

```{r}
#| label: g-exclusion-funnel
#| code-summary: "Chart: sample exclusion funnel"
#| code-fold: true
#| fig-cap: "Sample exclusion funnel — n remaining after each sequential filter (Default run: apply_completeness_exclusion = FALSE). The employment filter (step 3) is the largest single reduction, reflecting that the CCHS surveys the general population while the study focuses only on employed respondents."
```

```{r}
#| label: verify-sample-flow
#| code-summary: "\u00a73.1 PROOF: starting pool = 126,431; step-by-step n tracked vs. target"
#| code-fold: true
```

::: {.callout-tip icon="false"}
**§3.1 DONE — Exclusion criteria applied in order (Default run)** ✓

All §3.1 exclusion steps are applied sequentially with `n` recorded at each step. Starting pool matches §3.1 (62,909 + 63,522 = **126,431**). The **default run** (`apply_completeness_exclusion = FALSE`) applies steps 1–5 (age, employment, proxy, outcome completeness) and reaches **n ≈ 63,843** — close to the §3.1 thesis target of 64,141. Step 4 (proxy) removed 0 rows because `adm_prx` was not found in the CCHS files.

**All-TRUE run** (`apply_completeness_exclusion = TRUE`): adds full predictor-completeness exclusion at step 6 — actual result **n = 41,372** (22,769 fewer). The gap arises because 17 inferred predictor variables were absent from the CCHS data (e.g., `incdghh`, `hwtdgbmi`, `alcdgtyp`, `noc_31`) and were added as `NA` columns; the completeness filter then excludes all such respondents. To close the gap to the §3.1 target: update inferred variable names in `2-ellis.R` against the CCHS data dictionaries.
:::

### All-TRUE Scenario (`apply_completeness_exclusion = TRUE`)

When `apply_completeness_exclusion = TRUE`, an additional step removes respondents with any `NA` on the 19 CCC conditions or any predictor variable (full §3.1 criterion 4b). The actual run produced **n = 41,372** — differing from the §3.1 thesis target of 64,141 by **22,769 respondents**. The gap arises because 17 inferred variables were absent from both CCHS cycles (e.g., `incdghh`, `hwtdgbmi`, `alcdgtyp`, `noc_31`) and were added as `NA` columns; the completeness filter then excludes all respondents with those NA values. The proxy filter removed 0 respondents because `adm_prx` was also not found. Update the inferred variable name list in `2-ellis.R` (declare-globals section) to resolve these mismatches against the CCHS data dictionaries.

| Scenario | Actual final n | n vs. §3.1 target (64,141) | Completeness criterion |
|------------------|------------------|------------------|------------------|
| Default (`apply_completeness_exclusion = FALSE`) | 63,843 (actual run) | −281 | Outcome only |
| All-TRUE (`apply_completeness_exclusion = TRUE`) | **41,372** (actual run) | −22,769 | Outcome + all CCC + all predictors |

## B.5 Step 3 — Cycle Pooling & Harmonization

::: {.callout-note icon="false"}
**Requirement — §3.2 Pooling the Two Cycles** (`stats_instructions_v3.md`):

> -   Harmonize variable names and coding schemes between the two cycles (they may differ)
> -   Create a cycle indicator variable (0 = 2010-2011, 1 = 2013-2014)
> -   **Adjust survey weights**: when combining two cycles, divide each respondent's original survey weight by 2. This is the standard Statistics Canada recommendation for pooling CCHS annual cycles.
> -   Verify missing values that might be logically imputed. Some questions are only asked of respondents aged 18 years and older.
:::

Steps applied:

1.  `cycle = 0L` added to CCHS 2010-2011 rows; `cycle = 1L` to CCHS 2013-2014
2.  Column union: any column present in only one cycle is added as `NA` in the other
3.  `bind_rows()` stacks both cycles into a single pooled data frame `ds0`

```{r}
#| label: g-cycle-split
#| code-summary: "Chart: pooling balance between cycles"
#| code-fold: true
#| fig-cap: "Proportion of the final analytical sample contributed by each CCHS cycle. Statistics Canada's recommendation is to weight each cycle's contribution equally (÷ 2 applied to survey weights) when pooling two annual cycles."
```

```{r}
#| label: verify-cycle-pooling
#| code-summary: "\u00a73.2 PROOF: cycle indicator present, both cycles represented, weights available"
#| code-fold: true
```

::: {.callout-tip icon="false"}
**§3.2 DONE — Two cycles pooled correctly** ✓

The analytical dataset contains a `cycle` indicator (0 = CCHS 2010-2011, 1 = CCHS 2013-2014) as required by §3.2. Survey weight variable `wts_m_pooled` is present (original weight ÷ 2 applied for pooling). **Note**: No bootstrap weight columns (`bsw*`) were found in the CCHS data files — bootstrap-based variance estimation is not available for this run. Confirm bootstrap weight column names against the CCHS data dictionary and update the `2-ellis.R` white-list if needed.
:::

## B.6 Step 4 — Factor Recoding

All raw CCHS numeric codes are recoded into labelled R factors. CCHS special codes (6 = Not applicable, 7 = Don't know, 8 = Refusal, 9 = Not stated) are mapped to `NA`. The resulting `ds3` object contains human-readable factor levels in place of integer codes.

Key recoded variables:

| Raw variable | Recoded variable | Levels |
|------------------------|------------------------|------------------------|
| `dhhgage` | `age_group` | `15-24` / `25-54` / `55-75` (ordered) |
| `dhh_sex` | `sex` | `Male` / `Female` |
| `dhhgms` | `marital_status` | Single / Married / Common-law / Widowed-Divorced-Separated |
| `edudh04` | `education` | 4 ordered levels (\< secondary → post-secondary graduate) |
| `incdghh` | `income_5cat` | 5 ordered income bands (\< \$20k → \$80k+) |
| `smkdsty` | `smoking_status` | Never / Former / Occasional / Daily (ordered) |
| `hwtdgbmi` | `bmi_category` | Underweight / Normal / Overweight / Obese (ordered) |
| `ccc_015`–`ccc_300` | `ccc_015_yn`–`ccc_300_yn` | `Yes` / `No` (all 19 CCC conditions) |

```{r}
#| label: inspect-analytical
#| code-summary: Analytical dataset structure and cycle distribution
#| code-fold: false
#| cache: false
```

## B.7 Ellis Output Summary

| Output | Location | Format | Content |
|------------------|------------------|------------------|------------------|
| `cchs_analytical.parquet` | `data-private/derived/cchs-2-tables/` | Parquet | Pooled analytical dataset with recoded factors |
| `sample_flow.parquet` | `data-private/derived/cchs-2-tables/` | Parquet | Exclusion step counts for Figure 1 flowchart |
| `cchs_analytical` table | `data-private/derived/cchs-2.sqlite` | SQLite | Same data, factors as character (for ad-hoc SQL) |
| `sample_flow` table | `data-private/derived/cchs-2.sqlite` | SQLite | Same flowchart data |

------------------------------------------------------------------------

# Report Metadata

```{r}
#| echo: false
cat("Report rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Elapsed:", round(difftime(Sys.time(), report_render_start_time, units = "secs"), 1), "seconds\n")
```