---
title: "EDA 4 — Section 4.1 (Alternative Take): LOP Component Decomposition"
subtitle: "eda-4 — Dissecting the Primary Outcome Variable into Its 8 Reason Categories"
author: "Analysis Team"
date: "2026-03-22"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: true
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This report provides an **alternative take on Section 4.1** (Outcome Variable Construction) of `stats_instructions_v3.md`. Where eda-3 compared the primary outcome (`days_absent_total`) to the sensitivity outcome (`days_absent_chronic`) and reported the required distributional statistics, this EDA opens the "black box" of the total by examining the 8 individual LOP reason categories from which it is constructed.

**Research Questions:**

1.  What is the weighted prevalence of each of the 8 LOP reason categories (what proportion of workers report ≥1 absent day for each reason)?
2.  How much does each LOP component contribute to the weighted mean of `days_absent_total` in absolute and relative terms?
3.  How many reason categories does a typical respondent report simultaneously, and what is the co-occurrence pattern across reason pairs?

Following the dual-file development pattern, all analytical code lives in `analysis/eda-4/eda-4.R`. This document sources those chunks for presentation.

# Definition of Terms

**LOP module**: Loss of Productivity module from the CCHS. Respondents are asked how many workdays they missed in the past 3 months due to each of 8 distinct health-related reasons.

**days_absent_total**: The primary outcome. Row-wise sum of all 8 LOP component variables (`lopg040` + `lopg070` + `lopg082` through `lopg086` + `lopg100`), capped at 90 days.

**LOP reason categories (8 components):**

| Variable  | Reason                                |
|-----------|---------------------------------------|
| `lopg040` | Own chronic condition                 |
| `lopg070` | Injury                                |
| `lopg082` | Cold                                  |
| `lopg083` | Flu / influenza                       |
| `lopg084` | Gastroenteritis (stomach flu)         |
| `lopg085` | Respiratory infection                 |
| `lopg086` | Other infectious disease              |
| `lopg100` | Other physical / mental health reason |

```{r}
#| label: knitr-root
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../')
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  out.width    = "960px",
  fig.width    = 10,
  fig.height   = 6,
  fig.path     = 'figure-png-iso/'
)
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-4/eda-4.R")
```

# Environment

```{r}
#| label: load-packages
#| code-summary: Packages used in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Project-level utilities
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Paths, LOP component map, constants
#| echo: true
#| results: show
#| code-fold: true
```

```{r}
#| label: load-data
#| code-summary: Load cchs_analytical.parquet
#| echo: true
#| cache: true
#| results: show
```

# Data Context

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: Tables used in this EDA
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: data-context-person
#| code-summary: Representative individual — all LOP components visible
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: data-context-distributions
#| code-summary: Non-missing counts and raw prevalence per component
#| echo: true
#| results: show
```

```{r}
#| label: tweak-data-0
#| code-summary: Reshape to long-form LOP dataset
#| echo: true
#| cache: true
#| results: hide
```

# Analysis

## G0 — Zero vs Non-Zero Absenteeism

**Framing question**: How does the analytical sample split between workers who reported no missed days and those who missed at least one?

```{r}
#| label: g0-data-prep
#| code-summary: Compute zero / non-zero group counts and weighted shares
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: g0
#| code-summary: Horizontal stacked bar — zero vs ≥1 absent day
#| echo: true
#| cache: true
#| fig-width: 10
#| fig-height: 3
#| fig-cap: "Horizontal stacked bar showing the split between workers with no missed days and those with ≥1 missed day. Each segment shows the unweighted count (n) and unweighted percentage share."
```

```{r}
#| label: g01-data-prep
#| code-summary: Restrict to ≥1-day group and assign shared day-range colour palette
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: g01
#| code-summary: Histogram — total days missed, bars coloured by day-range category
#| echo: true
#| cache: true
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Frequency histogram of total missed days for respondents reporting ≥1 day. Bar colour maps to the same five day-range categories used in G02, making the two graphs visually consistent."
```

```{r}
#| label: g02-data-prep
#| code-summary: Summarise ≥1-day group by day-range category
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: g02
#| code-summary: Horizontal stacked bar — day-range breakdown (1–5, 6–10, 11–15, 16–30, 31+)
#| echo: true
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Horizontal stacked bar showing the proportion of the ≥1-day group in each day-range band (1–5, 6–10, 11–15, 16–30, 31+). Colours match the histogram in G01."
#| cache: false
```

## G1 — Prevalence of Each LOP Reason Category

**Research Question 1**: What proportion of workers missed ≥1 workday for each of the 8 reason categories?

```{r}
#| label: g1-data-prep
#| code-summary: Compute weighted prevalence per reason
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: g1
#| code-summary: Horizontal bar chart — weighted prevalence, ordered
#| echo: true
#| cache: true
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Weighted % of respondents reporting ≥1 absent day per LOP reason. Chronic condition and 'other physical/mental health' dominate."
```

```{r}
#| label: g11
#| code-summary: Prevalence by survey cycle (2011 vs 2014)
#| echo: true
#| cache: true
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Cycle comparison. Differences between waves can indicate question-wording changes or real temporal shifts."
```

## G2 — Contribution of Each Component to the Total Outcome

**Research Question 2**: How much does each LOP reason contribute to the weighted mean of `days_absent_total`?

```{r}
#| label: g2-data-prep
#| code-summary: Weighted mean days per reason category
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: g2
#| code-summary: Dot-strip plot — absolute mean days per reason with % share
#| echo: true
#| cache: true
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Each dot shows the weighted mean days absent attributed to that reason across all respondents. Chronic condition typically contributes the largest share."
```

```{r}
#| label: g21
#| code-summary: Stacked bar — relative shares of component sum
#| echo: true
#| cache: true
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Relative share of each LOP reason in the total absent-day burden (weighted means). Useful for seeing which reasons dominate the composite outcome."
```

## G3 — Co-occurrence of Reason Categories

**Research Question 3**: How many categories does a typical respondent report, and which reason pairs co-occur most often?

```{r}
#| label: g3-data-prep
#| code-summary: Compute per-respondent reason counts and pairwise co-occurrence matrix
#| echo: true
#| cache: true
#| results: show
```

```{r}
#| label: g3
#| code-summary: Histogram — number of reason categories per respondent
#| echo: true
#| cache: true
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Most respondents report 0 or 1 reason categories. The long tail informs whether the additive construction of the total outcome is dominated by single-reason reporters."
```

```{r}
#| label: g31
#| code-summary: Pairwise co-occurrence heatmap
#| echo: true
#| cache: true
#| fig-width: 8.5
#| fig-height: 5.5
#| fig-cap: "Cell values show % of respondents reporting ≥1 day for both reasons in the row/column pair. Diagonal entries equal univariate prevalence."
```

# Session Info

```{r}
#| label: session-info
#| code-summary: Reproducibility record
#| echo: true
#| results: show
#| code-fold: true
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
sessionInfo()
```