---
title: "EDA 4 - Questions 5.1 and 5.2"
subtitle: "Missing Data Analysis and Descriptive Statistics (Table 1)"
author: "Oleksandr Koval"
date: 2026-03-24
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: true
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

This document **observes and documents** question **5.1** (Handling Missing Data) and question **5.2** (Descriptive Statistics — Table 1) from `stats_instructions_v3.md`, using the Ellis output in `data-private/derived/cchs-2-tables/`.

The analysis addresses:

-   **Section 5.1**: Apply Little's MCAR test to assess missingness patterns; document proportion of missing values per variable; visualize missingness using `naniar` functions.
-   **Section 5.2**: Produce comprehensive descriptive statistics (Table 1) for the overall pooled sample and stratified by CCHS cycle (2011 vs. 2014), including unweighted and weighted frequencies, proportions, and outcome variable means/SDs.

**Data genealogy for this report:**

```         
1-ferry.R  ──► cchs-1.sqlite ──► 2-ellis.R ──► cchs_analytical.parquet ──► EDA-4 (Q5.1, Q5.2)
```

**Research Questions:**

1.  Does the data satisfy the Missing Completely At Random (MCAR) assumption? What are the key missingness patterns?
2.  What proportion of each predictor variable is missing, and how should missing data be handled?
3.  What are the demographic, behavioral, and health characteristics of the pooled sample and by cycle?
4.  How do weighted and unweighted distributions differ, and what is the impact of survey weighting?

```{r}
#| label: knitr-root
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = '../../')
```

```{r}
#| label: set_options
#| code-summary: Report-wide formatting options
#| echo: false
#| results: hide
#| message: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results = 'show',
  message = FALSE,
  warning = FALSE,
  comment = NA,
  tidy = FALSE,
  out.width = "960px",
  fig.width = 10,
  fig.height = 6,
  fig.path = 'figure-png-iso/'
)
options(width = 100)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/eda-4/eda-4.R")
```

# Environment

> Reviews the components of the working environment. Non-technical readers may skip this section.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

## Data Sources

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

This analysis uses the Ellis output: `cchs_analytical.parquet` from `data-private/derived/cchs-2-tables/`. The table contains person-level records with survey weights, outcome variables (days absent), and predictor variables across CCHS cycles 2011 and 2014.

```{r}
#| label: load-data
#| code-summary: Load analytical data from Ellis parquet output
#| warning: false
#| code-fold: true
#| cache: true
```

```{r}
#| label: inspect-data-0
#| code-summary: Data overview and structure
#| code-fold: false
#| eval: true
#| cache: true
```

## Data Context

This section describes the data as it appears in this specific analysis.

```{r}
#| label: inspect-data-1
#| code-summary: Grain verification — confirm unit of analysis
#| code-fold: false
#| eval: true
#| cache: true
```

```{r}
#| label: data-context-tables
#| code-summary: Ellis tables used and rationale
#| code-fold: true
#| cache: true
print(data_context_tables)
print(data_context_grain)
```

```{r}
#| label: data-context-person
#| code-summary: Single-person record view
#| code-fold: true
#| cache: true
print(data_context_person)
```

```{r}
#| label: data-context-distributions
#| code-summary: Key variable distributions for this analysis
#| code-fold: true
#| cache: true
print(data_context_distributions)
```

# Analysis

## Part A — Question 5.1 Handling Missing Data {#part-a}

::: {.callout-note icon="false"}
**Requirement — §5.1** (`stats_instructions_v3.md`, verbatim):

Apply Little's MCAR test to assess the pattern of missingness across predictor variables.

-   Document the proportion of missing values per variable
-   Consider multiple imputation if appropriate, or treat missing values as a separate category if the proportion is small (\< 5%) and the MCAR assumption holds

Suggested R package for missingness visualization: `naniar` (functions `vis_miss()`, `gg_miss_var()`, `gg_miss_upset()`).
:::

This section addresses question 5.1 by testing MCAR assumptions, documenting variable-level missingness, and visualizing missingness patterns with interactive Plotly graphics saved to `analysis/eda-4/prints/`.

### Graph Family g1: Missing Data Patterns — Heatmap and Upset Plot

Research Question: What is the pattern of missingness across predictor variables?

```{r}
#| label: g1-data-prep
#| code-summary: Prepare data for missing data visualization
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: g1
#| code-summary: Missing data heatmap (Plotly)
#| eval: true
#| cache: false
```

```{r}
#| label: g1-display
#| code-summary: Render missingness heatmap (interactive)
#| cache: false
g1_missing_heatmap
```

### Graph Family g2: Missing Data by Variable

Research Question: What proportion of each variable is missing?

```{r}
#| label: g2
#| code-summary: Missing data by variable (Plotly)
#| eval: true
#| cache: false
```

```{r}
#| label: g2-display
#| code-summary: Render missingness-by-variable chart (interactive)
#| cache: false
g2_missing_by_var
```

```{r}
#| label: g21
#| code-summary: Missingness co-occurrence patterns (Plotly)
#| eval: true
#| cache: false
```

```{r}
#| label: g21-display
#| code-summary: Render missingness co-occurrence chart (interactive)
#| cache: false
g21_missing_upset
```

```{r}
#| label: g22
#| code-summary: Distribution of variable-level missingness (Plotly)
#| eval: true
#| cache: false
```

```{r}
#| label: g22-display
#| code-summary: Render variable-level missingness distribution (interactive)
#| cache: false
g22_missing_density
```

### Table t3: Little's MCAR Test and Missing Data Summary

Research Question: Does the data satisfy MCAR assumption? What are the missing data patterns?

```{r}
#| label: t3-data-prep
#| code-summary: Prepare predictors for Little's MCAR test
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: t3
#| code-summary: Little's MCAR test and clear missing-data summary tables
#| eval: true
#| cache: true
print(t3_mcar_pretty)
print(t3_decision)
t3_missing_table_clear %>%
  slice_head(n = 25) %>%
  knitr::kable(caption = "Q5.1 Missingness summary (top variables) with <5% rule flag")
```

```{r}
#| label: verify-q51-proof
#| code-summary: "§5.1 PROOF: MCAR result + missingness decision diagnostics"
#| cache: true
q5_1_proof_table <- t3_mcar_pretty %>%
  mutate(
    mcar_holds = ifelse(!is.na(p_value) && p_value >= 0.05, "Yes", "No")
  )

q5_1_small_missing_proof <- missing_summary_decision %>%
  transmute(
    n_variables,
    n_vars_missing,
    max_pct_missing = scales::percent(max_pct_missing, accuracy = 0.1),
    median_pct_missing = scales::percent(median_pct_missing, accuracy = 0.1),
    all_lt_5pct
  )

print(q5_1_proof_table)
print(q5_1_small_missing_proof)
print(t3_decision)
```

### Final MCAR Decision Table (end of §5.1)

```{r}
#| label: q51-final-mcar-table
#| code-summary: "Final clear MCAR table with p-value and decision note"
#| cache: true
t3_mcar_pretty %>%
  dplyr::select(test, statistic, df, p_value, interpretation) %>%
  knitr::kable(caption = "Final MCAR test result (Little's test): p-value and interpretation")

t3_decision %>%
  knitr::kable(caption = "Missing-data decision notes based on MCAR test and <5% rule")
```

::: {.callout-tip icon="false"}
**§5.1 DONE — Missing-data requirements completed** ✓

The report now includes: (i) Little's MCAR test output, (ii) variable-level missingness proportions, (iii) explicit decision logic for `<5%` missingness handling, and (iv) Plotly-based visuals saved in `analysis/eda-4/prints/`.
:::

------------------------------------------------------------------------

## Part B — Question 5.2 Descriptive Statistics (Table 1) {#part-b}

::: {.callout-note icon="false"}
**Requirement — §5.2** (`stats_instructions_v3.md`, verbatim):

Produce the following for the overall pooled sample, stratified by CCHS cycle (2011 vs. 2014):

-   Unweighted frequencies and proportions for all categorical predictor variables
-   Weighted frequencies and proportions (using survey weights)
-   Weighted mean and standard deviation of the outcome variable (days absent)

Use the `survey` package in R to correctly account for the complex sampling design.
:::

This section addresses question 5.2 with reader-friendly descriptive tables and additional visuals to clarify weighting effects and cycle-level outcome differences.

### Table t1: Descriptive Statistics (Overall and Stratified by Cycle)

Research Question: What are the demographic, behavioral, and health characteristics of the sample?

```{r}
#| label: t1-data-prep
#| code-summary: Prepare categorical and continuous variables for Table 1
#| code-fold: true
#| eval: true
#| cache: true
```

```{r}
#| label: t1
#| code-summary: Table 1 — Descriptive statistics (unweighted and weighted) stratified by CCHS cycle
#| eval: true
#| cache: true
t1_table1_preview %>%
  dplyr::arrange(variable, group, level) %>%
  knitr::kable(caption = "Table 1: Categorical predictors (unweighted and weighted frequencies/proportions)")

t1_table1_proof %>%
  knitr::kable(caption = "Table 1 proof: proportions sum to 100% within each variable-group")

t1_table1_comprehensive %>%
  dplyr::arrange(variable, group, level) %>%
  knitr::kable(caption = "Q5.2 comprehensive descriptive table: categorical variables (overall + by cycle; unweighted and weighted)")
```

### Table t2: Outcome Variable Descriptive Statistics

Research Question: What are the outcomes (days absent) across the sample and by cycle?

```{r}
#| label: t2-data-prep
#| code-summary: Prepare outcome variables for descriptive statistics
#| code-fold: true
#| eval: true
#| cache: false
```

```{r}
#| label: t2
#| code-summary: Outcome descriptive statistics — weighted mean and SD by cycle
#| eval: true
#| cache: true
t2_outcome_primary_pretty %>%
  knitr::kable(caption = "Q5.2 outcome summary (days_absent_total): unweighted n/mean/SD and weighted mean/SE/SD")

t1_outcome_comprehensive %>%
  knitr::kable(caption = "Q5.2 comprehensive outcome descriptive table (overall + by cycle)")
```

### Static ggplot copies saved to `prints/`

In addition to interactive Plotly visuals, default `ggplot2` PNG copies are saved in `analysis/eda-4/prints/` for each graph family (`g1`, `g2`, `g21`, `g22`, `g3`, `g31`).

### Graph Family g3: Q5.2 Supporting Visuals

```{r}
#| label: g3
#| code-summary: Build weighted mean outcome graph (Plotly)
#| eval: true
#| cache: false
```

```{r}
#| label: g3-display
#| code-summary: Render weighted mean outcome graph (interactive)
#| cache: false
g3_outcome_weighted_mean_plot
```

```{r}
#| label: g31
#| code-summary: Build weighting-impact comparison graph (Plotly)
#| eval: true
#| cache: false
```

```{r}
#| label: g31-display
#| code-summary: Render weighting-impact comparison graph (interactive)
#| cache: false
g31_weighting_impact_plot
```

```{r}
#| label: verify-q52-proof
#| code-summary: "§5.2 PROOF: requirements checklist and survey-design evidence"
#| cache: true
print(t5_2_requirement_proof)

t5_2_requirement_proof %>%
  count(status) %>%
  knitr::kable(caption = "Q5.2 requirement checklist status")
```

::: {.callout-tip icon="false"}
**§5.2 DONE — Descriptive requirements completed** ✓

The report now provides unweighted and weighted categorical distributions, weighted outcome mean/SD, cycle stratification, and explicit proof checks that the required outputs were produced using a survey design object.
:::

------------------------------------------------------------------------

## Report Timeline

Report started: `r report_render_start_time`

Report rendered: `r Sys.time()`

Total time: `r round(difftime(Sys.time(), report_render_start_time, units = "secs"), 2)` seconds