---
title: "Table Anatomy: A Structural First Look at cchs_analytical"
subtitle: "Binder 2 — Notebook 1 · CCHS Pooled 2010/11 + 2013/14"
author: "Research Analyst"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

Before any variable is analysed, this notebook answers the most basic question a
tidy-trained analyst should ask: **what exactly is `cchs_analytical`?**

We walk through:

- How many rows and columns does the pooled table contain, and what domains do they represent?
- Which variables are fully available, and which are suppressed in the PUMF (public-use microdata)?
- How are the two survey cycles distributed — unweighted and weighted?
- What R types are used for each analytical domain, and do ordered factors carry the right structure?

The code follows the **graph family protocol**: each `gN-data-prep` chunk builds a shared
ancestor dataset; `gN` and `gN1`, `gN2`… draw multiple perspectives from the same source.

# Environment

```{r}
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = normalizePath("../../")) # must be its own chunk so root.dir is active for read_chunk()
```

```{r}
#| label: set-options
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = "show",
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/binder-2/1-table-anatomy.R")
```

```{r}
#| label: load-packages
#| code-summary: "Packages loaded for this notebook"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: "Project-level helper functions"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: "Paths, palette, and domain classification map"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: "classify_var_type() helper"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-data
#| code-summary: "Load cchs_analytical.parquet"
#| echo: true
#| results: show
#| message: true
#| code-fold: true
```

```{r}
#| label: tweak-data-0
#| echo: false
#| results: hide
```

# Data Context

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: "Table dimensions and cycle breakdown"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-person
#| code-summary: "Representative respondent — one row per cycle"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-distributions
#| code-summary: "Column completeness summary"
#| echo: true
#| code-fold: true
```

# Analysis

## Variable Completeness by Domain (g1 family)

`g1` maps *all 62 columns* simultaneously — grouping by domain and colouring by
domain so the eye can spot patterns within analytical zones.  
`g11` re-sorts the same data by completeness alone, pushing ALL-NA variables to
the bottom where they are impossible to miss.

```{r}
#| label: g1-data-prep
#| code-summary: "Build completeness + domain tibble for all 62 cols"
#| echo: true
#| message: false
#| code-fold: true
```

```{r}
#| label: g1
#| fig-cap: "Variable completeness map — 62 columns grouped by analytical domain. Bars at 0% are PUMF-suppressed."
#| fig-height: 11
#| echo: true
#| code-fold: true
print(g1_variable_map)
```

```{r}
#| label: g11
#| fig-cap: "Same data ranked by completeness — PUMF gaps (ALL-NA variables) cluster at the bottom."
#| fig-height: 11
#| echo: true
#| code-fold: true
print(g11_missing_ranked)
```

## Sample Size and Pooling Check (g2 family)

`g2` shows unweighted observation counts per cycle.  
`g21` compares weighted vs unweighted cycle shares to verify that the `wts_m / 2`
pooling rule was applied — both shares should land near 50%.

```{r}
#| label: g2-data-prep
#| code-summary: "Compute unweighted counts and weighted population sums by cycle"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g2
#| fig-cap: "Unweighted sample size by survey cycle."
#| fig-height: 5
#| echo: true
#| code-fold: true
print(g2_sample_by_cycle)
```

```{r}
#| label: g21
#| fig-cap: "Weighted vs unweighted cycle shares — both should be ~50%, confirming the halved-weight pooling rule."
#| fig-height: 5
#| echo: true
#| code-fold: true
print(g21_pooling_check)
```

## Variable Type Inventory (g3 family)

`g3` answers: for each analytical domain, what R types are used? Ordered factors are
critical for correct ordinal regression; this chart confirms their presence.

```{r}
#| label: g3-data-prep
#| code-summary: "Count variable types by analytical domain"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g3
#| fig-cap: "Stacked bar: variable types by analytical domain. Ordered factors indicate proper ordinal structure."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g3_type_inventory)
```

# Feedback Questions

1. **PUMF gaps**: The variable map immediately reveals 5+ columns that are ALL-NA
   (PUMF-suppressed). Which of these missing predictors would you consider most
   analytically consequential — and would you like a sensitivity analysis that
   models what those variables *might* have contributed?

2. **Domain classification**: The `domain_map` hard-codes assignments for 60 columns;
   2 fall through to "Other". Are there reclassifications you'd want (e.g., should
   `geodgprv` be "Survey Design" rather than "Facilitating")?

3. **Pooling verification**: `g21` shows both shares near 50%. If they were *not*
   equal, what would that imply about the weighting strategy, and what corrective
   step would you take before any weighted analysis?
