---
title: "Sociodemographic Fabric: Who Are the Workers in This Sample?"
subtitle: "Binder 2 — Notebook 2 · CCHS Pooled 2010/11 + 2013/14"
author: "Research Analyst"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

Statistical models are only as interpretable as our understanding of *who* is in the sample. This notebook paints a sociodemographic portrait of the 63,843 employed Canadians in `cchs_analytical`, answering:

-   What is the age, sex, education, income, immigration, and minority-status profile?
-   Are these profiles consistent across the two survey cycles?
-   How are employment type and work schedule distributed — and do they differ by sex?
-   Which provinces contribute the most respondents, and is provincial representation stable across cycles?

# Environment

```{r}
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = normalizePath("../../"))
```

```{r}
#| label: set-options
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = "show",
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/binder-2/2-sociodemographic-fabric.R")
```

```{r}
#| label: load-packages
#| code-summary: "Packages loaded for this notebook"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: "Project-level helper functions"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: "Paths, palette, variable lists, province lookup"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| echo: false
#| results: hide
```

```{r}
#| label: load-data
#| code-summary: "Load cchs_analytical.parquet"
#| echo: true
#| results: show
#| message: true
#| code-fold: true
```

```{r}
#| label: tweak-data-0
#| code-summary: "Add province label column from geodgprv codes"
#| echo: true
#| results: hide
#| code-fold: true
```

# Data Context

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: "Cycle breakdown table"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-person
#| code-summary: "Representative respondent — one row per cycle"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-distributions
#| code-summary: "Quick-look: age group and sex distributions"
#| echo: true
#| code-fold: true
```

# Analysis

## Six-Dimension Demographic Portrait (g1 family)

`g1` assembles six demographic variables into a single faceted view — one panel per variable, bars sorted by frequency. This is the fastest way to read the sample composition without flipping between six separate charts.

`g11` layers in the cycle dimension: parallel bars per level let the analyst check whether demographic proportions shifted between 2010/11 and 2013/14. Large discrepancies would warrant investigation before pooling.

```{r}
#| label: g1-data-prep
#| code-summary: "Pivot 6 demographic vars into long format; compute proportions"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g1
#| fig-cap: "Sociodemographic portrait of the pooled analytical sample. Six demographic variables displayed as faceted horizontal bars."
#| fig-height: 9
#| echo: true
#| code-fold: true
print(g1_demo_portrait)
```

```{r}
#| label: g11
#| fig-cap: "Demographic profile by survey cycle — bars coloured by cycle test for cross-cycle consistency."
#| fig-height: 9
#| echo: true
#| code-fold: true
print(g11_cycle_consistency)
```

## Employment Structure (g2 family)

`g2` cross-tabulates employment type (Employee / Self-employed / Unpaid) against work schedule (Full-time / Part-time) to reveal the dominant labour-market positions in the sample.

`g21` stacks employment types within sex, highlighting structural gender differences in self-employment and unpaid family work — both relevant potential moderators of absenteeism.

```{r}
#| label: g2-data-prep
#| code-summary: "Employment type × work schedule cross-tab"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g2
#| fig-cap: "Employment type by work schedule. Grouped bars show the full 3 × 2 cross-tab."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g2_employment_schedule)
```

```{r}
#| label: g21
#| fig-cap: "Employment type composition by sex. Stacked proportions reveal gender differences in self-employment."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g21_employment_by_sex)
```

## Geographic Distribution (g3 family)

`g3` ranks provinces by total respondent count — a quick read of where the analytical power lives.

`g31` checks cycle-to-cycle provincial balance: if provincial shares are stable across cycles, the pooled design is defensible. Substantial deviations would suggest either targeted oversampling in one cycle or changes in frame coverage.

```{r}
#| label: g3-data-prep
#| code-summary: "Province distribution, total and by cycle"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g3
#| fig-cap: "Sample size by province / territory, pooled. Sorted by total count."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g3_province)
```

```{r}
#| label: g31
#| fig-cap: "Provincial share within each cycle. Consistent proportions across cycles support pooling."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g31_province_cycle)
```

# Feedback Questions

1.  **Sole PUMF gap**: `student_status` is the only ALL-NA column in this dataset. Given that student status may interact with employment type and work schedule (part-time students vs. full-time workers), would you pursue an alternative operationalization from existing variables (e.g., age + education combination) or flag this as an unavoidable PUMF limitation?

2.  **Cycle consistency**: Looking at `g11`, is any demographic shift between 2010/11 and 2013/14 large enough to warrant a cycle-by-covariate interaction term in the regression model, or are the differences analytically negligible?

3.  **Employment type composition**: The sample contains only Employees (78%) and Self-employed (16%), with zero observations for "Unpaid family worker" (phantom PUMF level). Given this de facto binary split, should `employment_type` be simplified to a 2-level factor, or would you retain the original 3-level coding to accommodate potential future data linkage?