---
title: "Outcome Anatomy: Structure of days_absent_total"
subtitle: "Binder 2 — Notebook 3 · CCHS Pooled 2010/11 + 2013/14"
author: "Research Analyst"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

The primary outcome — `days_absent_total` — is a **count variable with extreme
zero-inflation**. Understanding its shape is not optional: it governs every
downstream modelling decision. This notebook examines:

- How severely zero-inflated is the outcome, and did the pattern change across cycles?
- Which of the eight LOP components contribute most to total absence, and are all
  components equally available in the data?
- What does the distribution of *non-zero* absences look like, and does it differ
  by age group?

Answers to these questions determine whether Negative Binomial, Zero-Inflated,
or Hurdle models are appropriate for the analysis.

# Environment

```{r}
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = normalizePath("../../"))
```

```{r}
#| label: set-options
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = "show",
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/binder-2/3-outcome-anatomy.R")
```

```{r}
#| label: load-packages
#| code-summary: "Packages loaded for this notebook"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: "Project-level helper functions"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: "Paths, palette, band levels, LOP variable names and labels"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| echo: false
#| results: hide
```

```{r}
#| label: load-data
#| code-summary: "Load cchs_analytical.parquet"
#| echo: true
#| results: show
#| message: true
#| code-fold: true
```

```{r}
#| label: tweak-data-0
#| code-summary: "Add absence_band column; confirm LOP variables present"
#| echo: true
#| results: show
#| message: true
#| code-fold: true
```

# Data Context

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: "Outcome descriptive statistics"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-person
#| code-summary: "Representative zero-absence and high-absence respondents"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-distributions
#| code-summary: "Outcome band distribution"
#| echo: true
#| code-fold: true
```

# Analysis

## Zero-Inflation Portrait (g1 family)

`g1` collapses the full 0–90 range into five interpretable bands and highlights
the structural zero cluster in red. The ~70% zero rate drives the entire modelling
strategy.

`g11` splits the same bands by survey cycle — if zero-inflation grew between
2010/11 and 2013/14, that would be a substantively meaningful trend worth
reporting.

```{r}
#| label: g1-data-prep
#| code-summary: "Outcome band counts, pooled and by cycle"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g1
#| fig-cap: "Zero-inflation portrait of days_absent_total. Red bar = structural zeros; blue bars = positive-count distribution."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g1_zero_inflation)
```

```{r}
#| label: g11
#| fig-cap: "Absence band distribution by survey cycle. Tests whether zero-inflation or positive tails shifted between cycles."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g11_by_cycle)
```

## LOP Component Decomposition (g2 family)

`days_absent_total` is the sum of eight LOP (Limitation of activity-related
work absences) components. `g2` shows how many days each component contributes
on average (among positive cases) and how completely it is measured.

`g21` shifts focus to data availability alone — some LOP components may be
more commonly answered than others, affecting the reliability of the composite.

```{r}
#| label: g2-data-prep
#| code-summary: "LOP component completeness and mean contribution"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g2
#| fig-cap: "Mean absence days per LOP component (positive cases). Fill = % non-missing. Longer bars indicate larger contribution to total absence."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g2_lop_components)
```

```{r}
#| label: g21
#| fig-cap: "LOP data availability — % non-missing per component. Components with low availability inflate missing rates in the composite."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g21_lop_completeness)
```

## Non-Zero Absence Distribution (g3 family)

`g3` focuses on the positive-count sub-distribution (respondents with at least
one absent day). Right skew, spikes at round numbers (5, 10, 15 days), and the
position of the median all inform model choice and dispersion assumptions.

`g31` facets this histogram by age group — a first bivariate signal that hints
at whether older workers tend toward more severe or more frequent absences.

```{r}
#| label: g3-data-prep
#| code-summary: "Restrict to positive-absence cases; compute median"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g3
#| fig-cap: "Distribution of work-absence days among positive-count respondents. Dashed line = median."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g3_nonzero_hist)
```

```{r}
#| label: g31
#| fig-cap: "Non-zero absence days faceted by age group. Reveals whether absence duration profiles differ across life stages."
#| fig-height: 7
#| echo: true
#| code-fold: true
print(g31_nonzero_by_age)
```

# Feedback Questions

1. **Model selection**: The zero-inflation rate (~70%) and the right-skewed
   positive count suggest Negative Binomial or a Hurdle model. Given that
   `days_absent_chronic` (the sensitivity outcome using only LOP040) is
   available, would you prefer to model the composite `days_absent_total`
   or would you run parallel analyses on both outcomes?

2. **LOP component granularity**: `g2` shows that some LOP components contribute
   more days than others on average. Should the analysis report component-specific
   results alongside the composite, or is the composite alone sufficient for the
   research question?

3. **Overdispersion check**: With mean ≈ median in the positive-count distribution
   but a very long right tail, variance almost certainly exceeds the mean. Would
   you like a formal overdispersion test (ratio of residual deviance to degrees
   of freedom from a Poisson GLM) added to this notebook before the modelling
   stage begins?
