---
title: "Raw to Analytical: Tracing Variable Transformations"
subtitle: "Binder 2 — Notebook 4 · CCHS Ferry (SQLite) → Ellis (Parquet)"
author: "Research Analyst"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
    keep-md: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Mission

Analytical results are only trustworthy if the data pipeline is transparent.
This notebook draws on **both raw and analytical data** to trace three variable
transformations end-to-end:

| Raw variable | Source table | Analytical variable | What Ellis does |
|---|---|---|---|
| `GENDHDI` | `cchs_2010_raw` / `cchs_2014_raw` | `self_health_general` | Recodes 6–9 → NA; applies ordered factor |
| `SMKDSCTY` | same raw tables | `smoking_status` | Recodes special codes → NA; reverses level order |
| `INCDGHH` (or equivalent) | same raw tables | `income_5cat` | Recodes refusals → NA; harmonizes across cycles |

For each variable the **g1/g11**, **g2/g21**, and **g3/g31** families show a
*before* chart (raw Ferry codes) and an *after* chart (analytical factor from
Ellis), making the transformation visible.

# Environment

```{r}
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir = normalizePath("../../"))
```

```{r}
#| label: set-options
#| echo: false
#| results: hide
#| message: false
#| warning: false
cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = "show",
  message    = FALSE,
  warning    = FALSE,
  comment    = NA,
  tidy       = FALSE,
  out.width  = "960px",
  fig.width  = 10,
  fig.height = 6,
  fig.path   = "figure-png-iso/"
)
ggplot2::theme_set(ggplot2::theme_minimal())
read_chunk("analysis/binder-2/4-raw-to-analytical.R")
```

```{r}
#| label: load-packages
#| code-summary: "Packages: arrow, DBI, RSQLite, ggplot2, dplyr, tidyr, forcats, scales"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: "Project-level helper functions"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-globals
#| code-summary: "Paths, palette, special-code list"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: "label_special() and raw_freq() helpers"
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-data
#| code-summary: "Load cchs_analytical.parquet + both raw SQLite tables"
#| echo: true
#| results: show
#| message: true
#| code-fold: true
```

```{r}
#| label: tweak-data-0
#| echo: false
#| results: hide
```

# Data Context

For complete data documentation, see the [Data Primer](../data-primer-1/data-primer-1.html).

```{r}
#| label: data-context-tables
#| code-summary: "Available raw variables — discovery check"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-person
#| code-summary: "Raw values for one respondent"
#| echo: true
#| code-fold: true
```

```{r}
#| label: data-context-distributions
#| code-summary: "Analytical distributions of self_health_general and smoking_status"
#| echo: true
#| code-fold: true
```

# Analysis

## Self-Rated Health: GENDHDI → self_health_general (g1 family)

CCHS uses a 1–5 scale (1 = Excellent … 5 = Poor) plus special codes 6, 7, 8,
9 for situations like refusal or "not applicable." Ellis recodes specials to NA
and applies an ordered factor with the correct level sequence.

`g1` shows the *raw* code frequency from the Ferry SQLite tables, with grey
bars identifying the special codes. `g11` shows the *analytical* distribution
after transformation — the factor levels properly ordered, specials gone.

```{r}
#| label: g1-data-prep
#| code-summary: "Tabulate GENDHDI raw codes by cycle from SQLite"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g1
#| fig-cap: "Before Ellis: raw GENDHDI frequency by cycle. Grey bars = special codes to be removed."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g1_raw_health)
```

```{r}
#| label: g11
#| fig-cap: "After Ellis: self_health_general ordered factor. Special codes removed; level ordering Excellent → Poor preserved."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g11_analytical_health)
```

## Smoking Status: SMKDSCTY → smoking_status (g2 family)

Raw SMKDSCTY codes follow a 1 (Daily) → 4 (Never) ordering that runs *counter*
to the natural "low exposure to high exposure" direction used in regression.
Ellis both removes special codes and reverses the level order, giving the
convention Never < Former < Occasional < Daily for model coefficients.

`g2` shows raw codes; `g21` shows the re-levelled analytical factor.

```{r}
#| label: g2-data-prep
#| code-summary: "Tabulate SMKDSCTY raw codes by cycle from SQLite"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g2
#| fig-cap: "Before Ellis: raw SMKDSCTY frequency by cycle. Grey bars = special codes."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g2_raw_smoking)
```

```{r}
#| label: g21
#| fig-cap: "After Ellis: smoking_status ordered factor (Never → Daily). Level order reversed relative to raw codes."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g21_analytical_smoking)
```

## Household Income: Raw codes → income_5cat (g3 family)

The income variable in the CCHS PUMF uses numeric codes that vary slightly
between cycles. Ellis identifies the correct variable, recodes refusal /
"not stated" codes to NA, and harmonizes both cycles into a shared 5-category
ordered factor.

The `data-context` chunk performs a live discovery of which raw income column
is present — the graph title reflects that finding at render time.

```{r}
#| label: g3-data-prep
#| code-summary: "Discover raw income variable; tabulate by cycle"
#| echo: true
#| code-fold: true
```

```{r}
#| label: g3
#| fig-cap: "Before Ellis: raw household income codes by cycle. Grey bars = refusal / not-stated codes."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g3_raw_income)
```

```{r}
#| label: g31
#| fig-cap: "After Ellis: income_5cat ordered factor. Cross-cycle harmonization and refusal removal completed."
#| fig-height: 5.5
#| echo: true
#| code-fold: true
print(g31_analytical_income)
```

# Feedback Questions

1. **Special-code magnitude**: Across the three variables, do the special-code
   rates (grey bars in the "Before" charts) reach a level where listwise deletion
   could introduce meaningful selection bias — and would multiple imputation
   be appropriate to assess whether the analytic estimates are sensitive to
   that exclusion?

2. **Cross-cycle harmonization**: `g31` shows the analytical income distribution
   by cycle after harmonization. Do you see substantial distributional shifts
   between 2010/11 and 2013/14 that suggest the raw income codes mapped to
   different real-income thresholds, and would you recommend reporting results
   with and without a cycle × income interaction?

3. **Traceability gap**: This notebook demonstrates distributional tracing
   (before/after marginal frequencies) but cannot show row-level tracing
   because the PUMF data do not include a persistent respondent ID.
   Would you consider adding a column-level crosswalk table (raw code → 
   analytical level) to the Ellis documentation to make the mapping fully
   auditable without row-level matching?
